---
title: "Control access to CerebroNexus with a login page"
author:
  - "Roman Hillje"
  - "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to CerebroNexus with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

Apps generated by `createShinyApp()` are public by default. To require a login,
provide an encrypted SQLite credentials database created by
[`shinymanager`](https://datastorm-open.github.io/shinymanager/) and keep its
passphrase in an environment variable.

The generated App contains the encrypted database, not the passphrase. Every
machine that runs the App must supply the same environment variable value.
Transport security remains the deployment operator's responsibility; serve a
login-protected App over HTTPS.

# Install the optional dependency

```{r eval=FALSE}
install.packages("shinymanager")
library(CerebroNexus)
```

# Create a multi-user database

For new deployments, use `provisionViewerAuthentication()` to create the
encrypted database, its one-line environment file, and the strict descriptor
consumed by `createShinyApp()`. This provisioning flow supports POSIX only and
fails closed on Windows. The handoff directory must be outside the App tree,
source control, web roots, sync folders, and ordinary backups. Plaintext login
passwords exist while this R call runs, so do not put real values in a committed
script or command history.

```{r eval=FALSE}
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = c(TRUE, FALSE),
  stringsAsFactors = FALSE
)

local({
  provision <- provisionViewerAuthentication(
    accounts = accounts,
    target_dir = "/srv/cerebro/private/viewer-auth"
  )
  readRenviron(provision$secret_file)
  on.exit(Sys.unsetenv(provision$auth$passphrase_env), add = TRUE)
  createShinyApp(
    cerebro_data = c(dataset = "dataset.crb"),
    result_dir = "/srv/cerebro/apps/viewer",
    auth = provision$auth
  )
})
```

The generated App contains only the encrypted database. Its separate
`viewer-auth.env` contains the passphrase and must remain outside the App tree.
Provisioning is deliberately local: it does not configure a server, container,
remote host, or secret store.

## Manual database creation (legacy)

Create a small environment file outside the App directory. Replace the example
value with a long random secret and restrict the file to its owner.

For example, OpenSSL can generate a 32-byte random value:

```bash
openssl rand -base64 32
```

Paste that output after `CEREBRO_AUTH_PASSPHRASE=` below. CerebroNexus rejects
passphrases shorter than 16 characters, but length alone is not a substitute
for randomness.

```{r eval=FALSE}
secret_file <- "my_app.env"
stopifnot(!file.exists(secret_file))
old_umask <- Sys.umask("077")
tryCatch(
  {
    stopifnot(file.create(secret_file))
    writeLines(
      "CEREBRO_AUTH_PASSPHRASE=replace-with-a-long-random-secret",
      secret_file
    )
  },
  finally = Sys.umask(old_umask)
)
Sys.chmod(secret_file, mode = "0600")
readRenviron(secret_file)
stopifnot(nzchar(Sys.getenv("CEREBRO_AUTH_PASSPHRASE")))
```

On Windows, POSIX file modes are not an access-control boundary. Store the
environment file in a location protected by the account's Windows ACLs, or use
the deployment platform's secret manager.

Create any number of users in one data frame. `shinymanager::create_db()` hashes
the login passwords and encrypts the credentials tables with the passphrase.

```{r eval=FALSE}
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = FALSE,
  stringsAsFactors = FALSE
)

shinymanager::create_db(
  credentials_data = accounts,
  sqlite_path = "credentials.sqlite",
  passphrase = Sys.getenv("CEREBRO_AUTH_PASSPHRASE")
)

rm(accounts)
```

`create_db()` hashes the passwords before storing them, and plaintext passwords
are not written to the generated App. During database creation they still exist
in the R data frame. Do not put real passwords in a committed script or retain
them in command history.

# Generate the App

```{r eval=FALSE}
createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_dataset.crb"),
  result_dir = "my_app",
  auth = list(
    credentials = "credentials.sqlite",
    passphrase_env = "CEREBRO_AUTH_PASSPHRASE",
    timeout_minutes = 15
  )
)
```

`createShinyApp()` verifies the database and passphrase before changing
`result_dir`, then copies only the encrypted database to
`my_app/private-data/auth/credentials.sqlite`.

# Run locally

A new R process does not inherit variables that were set in an earlier R
process. Load the environment file before starting the App:

```{r eval=FALSE}
readRenviron("/absolute/private/viewer-auth.env")
shiny::runApp("my_app")
```

Keep `viewer-auth.env` outside the App tree. Do not commit it or upload it with
the App directory. Unset the current-process secret when it is no longer needed:

```{r eval=FALSE}
Sys.unsetenv("CEREBRO_AUTH_PASSPHRASE")
```

# Limit failed sign-ins

Set a finite shinymanager failure policy in the R process that runs the App:

```{r eval=FALSE}
options("shinymanager.pwd_failure_limit" = 10L)
readRenviron("/absolute/private/viewer-auth.env")
shiny::runApp("my_app")
```

For Shiny Server, put the option in an `.Rprofile` read by its service account.
For a container, set it in the R command before `shiny::runApp()`. The counter
is stored in the encrypted database and locks an account after the configured
number of consecutive failures.

Account locking is defense in depth, not the only control: an attacker can use
it for targeted denial of service. Configure rate limiting at the reverse proxy
or ingress as well, preferably per source address, and monitor repeated failures.
Tune the proxy rule for Shiny's HTTP and WebSocket traffic rather than applying
an aggressive blanket limit to every Viewer request.

## Shiny Server or another R service

Load the file explicitly in the service bootstrap with
`readRenviron("/absolute/private/viewer-auth.env")` before the App starts.
Deploy only the generated App tree; keep the environment file outside the App
tree with service-account-only access. Provisioning does not change the service
configuration or deploy either controlled release member.

## systemd

Add `EnvironmentFile=/absolute/private/viewer-auth.env` to a separately
reviewed unit override, then restart through the operator's normal change
process. This package does not edit or reload the unit. The generated App tree
and its external secret handoff remain separate controlled release members.

## Docker or Docker Compose

Pass `--env-file /absolute/private/viewer-auth.env` through the deployment
system. Do not COPY the file into the image or mount it below the App tree.
Provisioning does not build, alter, or publish the container; review both the
App tree and external secret handoff as separate release members.

## Remote host or secret store

Transfer the handoff package to another machine over the organization's approved encrypted
channel, or import the one-line environment value into its secret store. Remove
the local handoff copy only after the service has been validated. Retain the old
release secret only for the approved rollback window, then revoke it and its
controlled backups. Provisioning does not contact remote hosts or secret stores;
the App tree and secret handoff remain separately controlled.

# Add or remove users

Update the source `credentials.sqlite` with shinymanager's database tools, then
run `createShinyApp()` again to publish a fresh encrypted copy. Keep the existing
passphrase, or rotate the database and deployed environment value together.

# See also

- [Create a self-contained Shiny App](create_a_self_contained_shiny_app.html)
- [Host CerebroNexus on shinyapps.io](host_cerebro_on_shinyapps.html)
- [`shinymanager::create_db()`](https://datastorm-open.github.io/shinymanager/reference/create_db.html)
