---
title: "Control access to CerebroNexus with a login page"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to CerebroNexus with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Security model

CerebroNexus supports opt-in Viewer authentication through `shinymanager`
(>= 1.1.0). Install that optional package before preparing credentials. When
`auth` is omitted or `NULL`, existing launches remain unauthenticated.

```{r install-provider, eval=FALSE}
install.packages("shinymanager")
```

The supported provider uses an encrypted SQLite credentials database.
`shinymanager` uses scrypt to protect each account password hash. Database
encryption is different: its key is derived directly as SHA-256(passphrase), so
a human-chosen password is unsuitable as the database passphrase. Generate at
least 32 bytes of high-entropy secret material with a cryptographically secure
random-number generator. Never print, paste into source, commit, or log that
value.

Authentication is only an application login gate. TLS/HTTPS is mandatory from
each browser to the deployment entry point and across every untrusted network
hop. The deployment platform or reverse proxy may terminate TLS, provided every
untrusted hop remains protected. The provider's login attempts are unlimited by
default. Put an Internet-facing app behind a reverse proxy or WAF for rate
limiting and request controls. Use an OIDC-capable access layer when
organizational SSO, MFA, centralized identity-provider revocation, or
public-access policy is required; the built-in provider does not supply those
controls.

# Create the credentials database

Create the database outside the source repository and outside every directory
served by Shiny or registered with `shiny::addResourcePath()`. The following
setup is deliberately non-evaluating: it must run only in a trusted interactive
administration session. It obtains the initial account name and password
interactively and contains no credential literals. The example uses `openssl`
for cryptographically secure random bytes and `askpass` for a non-echoing
password prompt; install those administration helpers if they are unavailable.

```{r install-admin-helpers, eval=FALSE}
install.packages(c("openssl", "askpass"))
```

```{r create-database, eval=FALSE}
create_initial_auth_database <- function(private_state_dir) {
  report_warning <- function(text) {
    tryCatch(
      warning(text, call. = FALSE),
      error = function(...) message("Warning: ", text)
    )
    invisible(NULL)
  }
  candidate <- NULL
  created_auth_dir <- FALSE
  published <- FALSE
  auth_dir <- NULL
  intended_auth_dir <- NULL
  created_auth_owner <- NA_integer_
  account_password <- NULL
  credentials <- NULL
  db_passphrase <- NULL
  candidate_tables <- NULL
  secret_name <- "CEREBRO_AUTH_PASSPHRASE"
  prior_secret <- Sys.getenv(secret_name, unset = NA_character_)
  prior_secret_was_set <- !is.na(prior_secret)
  secret_installed <- FALSE
  on.exit(
    {
      account_password <- NULL
      credentials <- NULL
      db_passphrase <- NULL
      candidate_tables <- NULL
      cleanup_incomplete <- FALSE
      if (
        is.character(candidate) &&
          length(candidate) == 1L &&
          !is.na(candidate)
      ) {
        candidate_artifacts <- c(
          candidate,
          paste0(candidate, c("-journal", "-wal", "-shm"))
        )
        existing_artifacts <- candidate_artifacts[
          file.exists(candidate_artifacts) |
            dir.exists(candidate_artifacts)
        ]
        removable_artifacts <- existing_artifacts[
          !dir.exists(existing_artifacts)
        ]
        cleanup_status <- if (length(removable_artifacts)) {
          vapply(
            removable_artifacts,
            unlink,
            integer(1),
            recursive = FALSE
          )
        } else {
          integer()
        }
        if (
          any(cleanup_status != 0L) ||
            any(
              file.exists(candidate_artifacts) |
                dir.exists(candidate_artifacts)
            )
        ) {
          cleanup_incomplete <- TRUE
        }
      }
      if (created_auth_dir && !published) {
        auth_link <- if (
          is.character(auth_dir) &&
            length(auth_dir) == 1L &&
            !is.na(auth_dir)
        ) {
          Sys.readlink(auth_dir)
        } else {
          NA_character_
        }
        auth_exists <- is.character(auth_dir) &&
          length(auth_dir) == 1L &&
          !is.na(auth_dir) &&
          (file.exists(auth_dir) || dir.exists(auth_dir))
        auth_is_link <- !is.na(auth_link) && nzchar(auth_link)
        if (auth_exists || auth_is_link) {
          resolved_cleanup_dir <- if (dir.exists(auth_dir) && !auth_is_link) {
            tryCatch(
              normalizePath(auth_dir, winslash = "/", mustWork = TRUE),
              error = function(...) NA_character_
            )
          } else {
            NA_character_
          }
          cleanup_dir_safe <- !auth_is_link &&
            !is.na(resolved_cleanup_dir) &&
            identical(resolved_cleanup_dir, intended_auth_dir)
          if (cleanup_dir_safe && .Platform$OS.type == "unix") {
            cleanup_owner <- file.info(auth_dir)$uid[[1L]]
            cleanup_dir_safe <- !is.na(cleanup_owner) &&
              !is.na(created_auth_owner) &&
              cleanup_owner == created_auth_owner
          }
          auth_contents <- if (cleanup_dir_safe) {
            tryCatch(
              list.files(
                auth_dir,
                all.files = TRUE,
                no.. = TRUE,
                recursive = FALSE
              ),
              error = function(...) NA_character_
            )
          } else {
            NA_character_
          }
          if (
            cleanup_dir_safe &&
              !anyNA(auth_contents) &&
              !length(auth_contents)
          ) {
            directory_removed <- suppressWarnings(file.remove(auth_dir))
            if (
              !isTRUE(directory_removed) ||
                file.exists(auth_dir) ||
                dir.exists(auth_dir)
            ) {
              cleanup_incomplete <- TRUE
            }
          } else {
            cleanup_incomplete <- TRUE
          }
        }
      }
      if (cleanup_incomplete) {
        report_warning(
          paste(
            "Unpublished authentication database cleanup was incomplete;",
            "manual recovery is required."
          )
        )
      }
      if (!published && secret_installed) {
        if (prior_secret_was_set) {
          do.call(
            Sys.setenv,
            stats::setNames(list(prior_secret), secret_name)
          )
        } else {
          Sys.unsetenv(secret_name)
        }
        restored_secret <- Sys.getenv(secret_name, unset = NA_character_)
        if (
          (prior_secret_was_set && !identical(restored_secret, prior_secret)) ||
            (!prior_secret_was_set && !is.na(restored_secret))
        ) {
          report_warning(
            "Authentication passphrase environment restoration failed; manual recovery is required."
          )
        }
      }
      prior_secret <- NULL
    },
    add = TRUE
  )

  if (
    !is.character(private_state_dir) ||
      length(private_state_dir) != 1L ||
      is.na(private_state_dir) ||
      !nzchar(private_state_dir)
  ) {
    stop("Private state directory must be one existing path.")
  }
  private_state_dir <- normalizePath(
    private_state_dir,
    winslash = "/",
    mustWork = TRUE
  )
  private_owner <- if (.Platform$OS.type == "unix") {
    file.info(private_state_dir)$uid[[1L]]
  } else {
    NA_integer_
  }
  auth_dir <- file.path(private_state_dir, "cerebro-auth")
  auth_link <- Sys.readlink(auth_dir)
  if (!is.na(auth_link) && nzchar(auth_link)) {
    stop("Authentication state directory must not be a symbolic link.")
  }
  if (file.exists(auth_dir) || dir.exists(auth_dir)) {
    stop("Authentication state directory must not already exist.")
  }
  if (!dir.create(auth_dir, mode = "0700")) {
    stop("Could not create the authentication state directory.")
  }
  created_auth_dir <- TRUE
  if (.Platform$OS.type == "unix") {
    created_auth_owner <- file.info(auth_dir)$uid[[1L]]
  }
  intended_auth_dir <- file.path(private_state_dir, "cerebro-auth")
  resolved_auth_dir <- normalizePath(
    auth_dir,
    winslash = "/",
    mustWork = TRUE
  )
  auth_link <- Sys.readlink(auth_dir)
  if (
    (!is.na(auth_link) && nzchar(auth_link)) ||
      !identical(resolved_auth_dir, intended_auth_dir)
  ) {
    stop("Authentication state directory resolved outside its intended location.")
  }
  auth_dir <- resolved_auth_dir
  if (.Platform$OS.type == "unix") {
    Sys.chmod(auth_dir, mode = "0700")
    auth_mode <- file.info(auth_dir)$mode[[1L]]
    auth_owner <- file.info(auth_dir)$uid[[1L]]
    if (
      is.na(auth_mode) ||
        as.integer(auth_mode) != as.integer(as.octmode("0700")) ||
        is.na(auth_owner) ||
        is.na(private_owner) ||
        auth_owner != private_owner
    ) {
      stop("Authentication state directory owner or mode could not be verified.")
    }
  }

  rechecked_auth_dir <- normalizePath(
    auth_dir,
    winslash = "/",
    mustWork = TRUE
  )
  auth_link <- Sys.readlink(auth_dir)
  if (
    (!is.na(auth_link) && nzchar(auth_link)) ||
      !identical(rechecked_auth_dir, intended_auth_dir)
  ) {
    stop("Authentication state directory changed before database creation.")
  }
  if (.Platform$OS.type == "unix") {
    rechecked_info <- file.info(auth_dir)
    if (
      is.na(rechecked_info$uid[[1L]]) ||
        rechecked_info$uid[[1L]] != auth_owner ||
        is.na(rechecked_info$mode[[1L]]) ||
        as.integer(rechecked_info$mode[[1L]]) !=
          as.integer(as.octmode("0700"))
    ) {
      stop("Authentication state directory changed before database creation.")
    }
  }

  credentials_path <- file.path(auth_dir, "credentials.sqlite")
  if (file.exists(credentials_path) || dir.exists(credentials_path)) {
    stop("The final credentials database already exists; refusing to overwrite it.")
  }
  repeat {
    candidate <- tempfile(
      pattern = ".credentials-",
      tmpdir = auth_dir,
      fileext = ".sqlite"
    )
    if (!file.exists(candidate) && !dir.exists(candidate)) {
      break
    }
  }

  account_name <- readline("Initial account name: ")
  if (
    !is.character(account_name) ||
      length(account_name) != 1L ||
      is.na(account_name) ||
      !nzchar(trimws(account_name))
  ) {
    stop("Initial account name must be non-empty.")
  }
  account_name <- trimws(account_name)
  account_password <- askpass::askpass("Initial account password")
  if (is.null(account_password)) {
    stop("Initial account password entry was cancelled.")
  }
  if (
    !is.character(account_password) ||
      length(account_password) != 1L ||
      is.na(account_password) ||
      !nzchar(account_password)
  ) {
    stop("Initial account password must be non-empty.")
  }
  db_passphrase <- openssl::base64_encode(openssl::rand_bytes(32))

  credentials <- data.frame(
    user = account_name,
    password = account_password,
    admin = FALSE,
    stringsAsFactors = FALSE
  )
  shinymanager::create_db(
    credentials_data = credentials,
    sqlite_path = candidate,
    passphrase = db_passphrase
  )

  candidate_tables <- list(
    credentials = shinymanager::read_db_decrypt(
      candidate,
      "credentials",
      db_passphrase
    ),
    pwd_mngt = shinymanager::read_db_decrypt(
      candidate,
      "pwd_mngt",
      db_passphrase
    ),
    logs = shinymanager::read_db_decrypt(
      candidate,
      "logs",
      db_passphrase
    )
  )
  required_columns <- list(
    credentials = c(
      "user",
      "password",
      "start",
      "expire",
      "admin",
      "is_hashed_password"
    ),
    pwd_mngt = c(
      "user",
      "must_change",
      "have_changed",
      "date_change",
      "n_wrong_pwd"
    ),
    logs = c("user", "server_connected", "token", "logout", "app")
  )
  tables_valid <- vapply(
    names(required_columns),
    function(name) {
      table <- candidate_tables[[name]]
      is.data.frame(table) &&
        !anyDuplicated(names(table)) &&
        all(required_columns[[name]] %in% names(table))
    },
    logical(1)
  )
  hashes <- candidate_tables$credentials$is_hashed_password
  if (
    !all(tables_valid) ||
      nrow(candidate_tables$credentials) != 1L ||
      length(hashes) != 1L ||
      is.na(hashes) ||
      hashes != 1
  ) {
    stop("Candidate credentials database validation failed.")
  }

  if (.Platform$OS.type == "unix") {
    Sys.chmod(candidate, mode = "0600")
    candidate_mode <- file.info(candidate)$mode[[1L]]
    candidate_owner <- file.info(candidate)$uid[[1L]]
    if (
      is.na(candidate_mode) ||
        as.integer(candidate_mode) != as.integer(as.octmode("0600")) ||
        is.na(candidate_owner) ||
        candidate_owner != auth_owner
    ) {
      stop("Candidate credentials database owner or mode is unsafe.")
    }
  }
  if (file.access(candidate, mode = 6L) != 0L) {
    stop("Candidate credentials database is not readable/writable.")
  }
  candidate_sidecars <- paste0(candidate, c("-journal", "-wal", "-shm"))
  if (any(file.exists(candidate_sidecars) | dir.exists(candidate_sidecars))) {
    stop("The candidate database still has active SQLite sidecars.")
  }
  if (file.exists(credentials_path) || dir.exists(credentials_path)) {
    stop("The final credentials database appeared during setup; refusing to overwrite it.")
  }
  secret_installed <- TRUE
  do.call(
    Sys.setenv,
    stats::setNames(list(db_passphrase), secret_name)
  )
  if (!identical(Sys.getenv(secret_name, unset = NA_character_), db_passphrase)) {
    stop("Could not install the authentication passphrase environment variable.")
  }
  if (!suppressWarnings(file.link(candidate, credentials_path))) {
    stop(
      paste(
        "The filesystem must support same-directory hard links for atomic",
        "no-clobber publication; use a platform atomic no-clobber mechanism.",
        "No rename fallback was attempted."
      )
    )
  }
  published <- TRUE
  if (
    unlink(candidate, recursive = FALSE) != 0L ||
      file.exists(candidate) ||
      dir.exists(candidate)
  ) {
    report_warning(
      sprintf(
        paste(
          "The database was published, but the published candidate remains at",
          "%s; remove that hard-link name manually."
        ),
        candidate
      )
    )
  }
  candidate <- NULL
  invisible(credentials_path)
}

credentials_path <- create_initial_auth_database(
  Sys.getenv("CEREBRO_PRIVATE_STATE_DIR")
)
```

The example enforces non-empty input as its minimum. Apply the organization's
password length, breach-list, and rotation policy before production use. The
function refuses to overwrite an existing final database. It publishes only a
validated owner-only sibling using a same-directory atomic no-clobber hard link,
then removes the candidate name. The deployment filesystem must support
same-directory hard links; an unsupported link or a raced target fails closed,
with no fallback to an overwriting rename. Use a deployment-platform atomic
no-clobber mechanism where hard links are unavailable. On failure the function
attempts to remove the candidate and its SQLite sidecars, then removes the
directory it created only after revalidating its path and owner and proving it
is empty. Cleanup is non-recursive, and any residue produces a manual-recovery
warning without hiding the original error. Before publication, the function
installs the new passphrase in `CEREBRO_AUTH_PASSPHRASE`; if publication fails,
it restores the exact prior value or unsets a previously absent variable. Once
the hard link succeeds, the database and matching environment secret are kept.
A failure to remove the now-redundant candidate name produces a warning with its
recovery path but does not turn the usable publication into an error. The
function drops its local references to secret-bearing objects on exit and
returns only the final database path.

For deployed services, inject `CEREBRO_AUTH_PASSPHRASE` with the platform's
secret manager. Never place it in an app configuration file, R source, a shell
command committed with the deployment, or logs. The application reads the
named environment variable when validating and opening the database; the
descriptor contains only the variable's name.

Protect the database file and every parent directory. The runtime identity needs
read/write permission on the file, write/search permission on its containing
directory to create SQLite journal files, and search permission on the remaining
parents. On POSIX systems, prefer an owner-only directory (`0700`) and
owner-readable/writable database (`0600`), owned by the service account.
Equivalent ACL and ownership controls are the deployment platform's
responsibility. Do not make either path web-readable to solve a permissions
error.

# Configure authentication

Both public entry points accept the same exact descriptor. `credentials` is an
absolute path, `passphrase_env` is the name of the injected environment
variable, and `timeout_minutes` is an optional whole number from 1 through 1440
(default 15). For a direct launch, the host database must be a regular
readable/writable file outside HTTP resources; its containing directory must be
writable/searchable for encrypted logs and SQLite journals, and other parents
must be searchable. During bundle creation, the source database need only be a
regular readable file. On the deployment host, the generated private copy must
be readable/writable and its containing directory writable/searchable.

```{r descriptor, eval=FALSE}
auth <- list(
  provider = "shinymanager",
  credentials = credentials_path,
  passphrase_env = "CEREBRO_AUTH_PASSPHRASE",
  timeout_minutes = 15
)
```

For a direct Viewer, pass the descriptor without changing CerebroNexus source:

```{r direct-launch, eval=FALSE}
CerebroNexus::launchCerebro(mode = "closed", auth = auth)
```

For generated bundle deployment, use the same descriptor:

```{r generated-bundle, eval=FALSE}
CerebroNexus::createShinyApp(cerebro_data, result_dir, auth = auth)
```

The generated bundle copies the validated database to the fixed private path
`private-data/auth/credentials.sqlite`; the passphrase is not copied. Configure
the same named secret on the deployment host. The host must preserve a writable
database file and a writable/searchable parent directory for SQLite journals.
The generated path is not registered as a Shiny resource, but filesystem
permissions remain mandatory.

# Login lifecycle and account changes

The Viewer server starts for a browser session only after the provider accepts
an active account from the database. `timeout_minutes` controls the provider's
session timeout. Selecting logout returns the user to the login page, and login
tokens are not persisted across sessions. Provider login logs are stored in the
same encrypted database, which is why runtime write access is required.

The in-app `shinymanager` administration UI is disabled. To add, remove, expire,
or reset accounts, prepare the complete replacement credentials data in a
trusted administration session and collect any new passwords interactively.
Stop the affected app instance and back up the encrypted database. Run
`shinymanager::create_db()` with the same deployment-managed passphrase but set
`sqlite_path` to a fresh sibling path in the database directory, not the live
path. Decrypt and validate that candidate, apply owner-only permissions, then
use the deployment platform's atomic replace operation to move the sibling over
the live path without a copy/delete window. Restart only after the replacement
succeeds. For a generated app, rebuild and redeploy the bundle so the fixed
private copy is updated; editing the source database does not modify an already
generated bundle.

# Process and deployment topology

Run one secured Viewer per R process. `shinymanager` and CerebroNexus maintain
process-global state that does not isolate two separately sourced secured apps
inside one process. In a multi-worker deployment, use sticky routing so a
browser session continues to reach the worker that owns its Shiny session. Each
worker must receive the same secret-manager binding and filesystem access.

# Migration from historical examples

The old hand-edited `secure_app()` workflow is unsupported: do not patch
`app.R`, `shiny_UI.R`, or `shiny_server.R`, and do not embed a credentials data
frame in source. The archived custom SHA-256 verifier is unsupported as well;
it lacks the maintained database, password-hashing, session, and validation
contract of the public `auth` interface. Migrate both approaches to the strict
descriptor shown above.

# See also

* [Launch Cerebro with a pre-loaded data set](launch_cerebro_with_pre-loaded_data_set.html)
* [Host Cerebro on shinyapps.io](host_cerebro_on_shinyapps.html)
