---
title: "Control access to CerebroNexus with a login page"
author:
  - "Roman Hillje"
  - "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to CerebroNexus with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

Apps generated by `createShinyApp()` are public by default. To require a login,
provide an encrypted SQLite credentials database created by
[`shinymanager`](https://datastorm-open.github.io/shinymanager/) and keep its
passphrase in an environment variable.

The generated App contains the encrypted database, not the passphrase. Every
machine that runs the App must supply the same environment variable value.
Transport security remains the deployment operator's responsibility; serve a
login-protected App over HTTPS.

# Install the optional dependency

```{r eval=FALSE}
install.packages("shinymanager")
library(CerebroNexus)
```

# Create a multi-user database

Create a small environment file outside the App directory. Replace the example
value with a long random secret and restrict the file to its owner.

```{r eval=FALSE}
secret_file <- "my_app.env"
stopifnot(!file.exists(secret_file))
old_umask <- Sys.umask("077")
tryCatch(
  {
    stopifnot(file.create(secret_file))
    writeLines(
      "CEREBRO_AUTH_PASSPHRASE=replace-with-a-long-random-secret",
      secret_file
    )
  },
  finally = Sys.umask(old_umask)
)
Sys.chmod(secret_file, mode = "0600")
readRenviron(secret_file)
stopifnot(nzchar(Sys.getenv("CEREBRO_AUTH_PASSPHRASE")))
```

On Windows, POSIX file modes are not an access-control boundary. Store the
environment file in a location protected by the account's Windows ACLs, or use
the deployment platform's secret manager.

Create any number of users in one data frame. `shinymanager::create_db()` hashes
the login passwords and encrypts the credentials tables with the passphrase.

```{r eval=FALSE}
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = FALSE,
  stringsAsFactors = FALSE
)

shinymanager::create_db(
  credentials_data = accounts,
  sqlite_path = "credentials.sqlite",
  passphrase = Sys.getenv("CEREBRO_AUTH_PASSPHRASE")
)

rm(accounts)
```

`create_db()` hashes the passwords before storing them, and plaintext passwords
are not written to the generated App. During database creation they still exist
in the R data frame. Do not put real passwords in a committed script or retain
them in command history.

# Generate the App

```{r eval=FALSE}
createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_dataset.crb"),
  result_dir = "my_app",
  auth = list(
    credentials = "credentials.sqlite",
    passphrase_env = "CEREBRO_AUTH_PASSPHRASE",
    timeout_minutes = 15
  )
)
```

`createShinyApp()` verifies the database and passphrase before changing
`result_dir`, then copies only the encrypted database to
`my_app/private-data/auth/credentials.sqlite`.

# Run locally

A new R process does not inherit variables that were set in an earlier R
process. Load the environment file before starting the App:

```{r eval=FALSE}
readRenviron("my_app.env")
shiny::runApp("my_app")
```

Keep `my_app.env` outside `my_app`. Do not commit it or upload it with the
App directory.

# Shiny Server and systemd

Install the generated App under the Shiny Server application directory and put
the secret in a root-managed file:

```bash
sudo install -d -m 0755 /srv/shiny-server/my_app
sudo cp -R my_app/. /srv/shiny-server/my_app/
sudo install -d -m 0750 /etc/cerebronexus
sudo install -o root -g shiny -m 0640 my_app.env /etc/cerebronexus/my_app.env
sudo chown -R shiny:shiny /srv/shiny-server/my_app/private-data/auth
sudo chmod 0700 /srv/shiny-server/my_app/private-data/auth
sudo chmod 0600 /srv/shiny-server/my_app/private-data/auth/credentials.sqlite
```

If Shiny Server is managed by systemd, add a service drop-in with
`sudo systemctl edit shiny-server`:

```ini
[Service]
EnvironmentFile=/etc/cerebronexus/my_app.env
```

Then restart the service:

```bash
sudo systemctl daemon-reload
sudo systemctl restart shiny-server
```

The service account needs read/write access to the bundled SQLite database and
its containing directory because shinymanager writes login logs and SQLite
journal files.

# Docker Compose

A minimal image installs the runtime packages and copies the generated App:

```dockerfile
FROM rocker/shiny:latest
RUN R -q -e 'install.packages("remotes"); remotes::install_github("mihem/CerebroNexus", dependencies=TRUE); install.packages("shinymanager")'
COPY my_app /srv/shiny-server/my_app
RUN chown -R shiny:shiny /srv/shiny-server/my_app
USER shiny
CMD ["R", "-q", "-e", "shiny::runApp('/srv/shiny-server/my_app', host='0.0.0.0', port=3838)"]
```

Docker Compose can load the key without baking it into the image:

```yaml
services:
  cerebro:
    build: .
    env_file:
      - ./my_app.env
    ports:
      - "3838:3838"
```

Keep `my_app.env` out of the Docker build context or exclude it in
`.dockerignore`.

# Run on another machine

Copy the `my_app` directory and the secret through separate protected channels.
On another machine, restore the same variable name and value before running the
App:

```{r eval=FALSE}
readRenviron("/protected/path/my_app.env")
shiny::runApp("/srv/cerebronexus/my_app")
```

If the environment value does not match the database, the App stops instead of
starting with an unusable login page.

# Add or remove users

Update the source `credentials.sqlite` with shinymanager's database tools, then
run `createShinyApp()` again to publish a fresh encrypted copy. Keep the existing
passphrase, or rotate the database and deployed environment value together.

# See also

- [Create a self-contained Shiny App](create_a_self_contained_shiny_app.html)
- [Host CerebroNexus on shinyapps.io](host_cerebro_on_shinyapps.html)
- [`shinymanager::create_db()`](https://datastorm-open.github.io/shinymanager/reference/create_db.html)
