---
title: "Control access to Cerebro with a login page"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to Cerebro with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

When hosting Cerebro on a web server, it can be useful to control who has access to it, especially, when it is pre-loaded with a data set (see `vignette('launch_cerebro_with_pre-loaded_data_set')`).
This can be done with the [`shinymanager`](https://cran.r-project.org/web/packages/shinymanager/index.html) package, however some (minor) adjustments must be made to the source code of Cerebro.

> **Where these files live.** The edits below assume a bundle produced by
> `createShinyApp()` (see
> [Create a self-contained Shiny app](create_a_self_contained_shiny_app.html)
> and [Host on shinyapps.io](host_cerebro_on_shinyapps.html)): `app.R` at the
> bundle root, `shiny_server.R`/`shiny_UI.R` under `shiny/v1.4/` inside it.

# Preparation

## `app.R`

At the top of the `app.R` file, add the `shinymanager` package to be loaded alongside other packages that are required by Cerebro:

```{r eval=FALSE}
## load packages -------------------------------------------------------------##
library(shinymanager)
library(dplyr)
library(DT)
# ...
```

## `shiny_server.R`

Add this to the top of the `shiny_server.R` file, before the definition of the `server` function starts:

```{r eval=FALSE}
user_credentials <- tibble::tribble(
   ~user,              ~password,       ~start, ~expire, ~admin,
  "user1", "reallygoodpassword1", "2020-08-01",      NA,  FALSE,
  "user2", "reallygoodpassword2", "2020-09-01",      NA,  FALSE
)
```

Within the definition of the `server` function, right at the top, add this:

```{r eval=FALSE}
res_auth <- shinymanager::secure_server(
  check_credentials = shinymanager::check_credentials(user_credentials)
)
```

The head of the `shiny_server.R` file should now look similar to this:

```{r eval=FALSE}
## define users that can access Cerebro --------------------------------------##
user_credentials <- tibble::tribble(
   ~user,              ~password,       ~start, ~expire, ~admin,
  "user1", "reallygoodpassword1", "2020-08-01",      NA,  FALSE,
  "user2", "reallygoodpassword2", "2020-09-01",      NA,  FALSE
)

## server function -----------------------------------------------------------##
server <- function(input, output, session) {
  
  ## check user authentication -----------------------------------------------##
  res_auth <- secure_server(
    check_credentials = check_credentials(user_credentials)
  )
# ...
```

## `shiny_UI.R`

At the end of the file, after the definition of the `ui` function, add this:

```{r eval=FALSE}
ui <- shinymanager::secure_app(ui)
```

# Usage

Now, when opening the Cerebro user interface, `shinymanager` shows a login
form before anything else loads; users have to authenticate themselves as one
of the users in the `user_credentials` data frame.
Of course, the use credentials can also be loaded from a file, which might be easier to manage.

Thanks to Michael Heming from the University of Münster for pointing this possibility out and sharing his implementation.

# See also

* [Launch Cerebro with pre-loaded data set](launch_cerebro_with_pre-loaded_data_set.html)
* [Host Cerebro on shinyapps.io](host_cerebro_on_shinyapps.html)
