---
title: "Control access to CerebroNexus with a login page"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to CerebroNexus with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`createShinyApp(auth = TRUE)` is the shortest supported way to build a
multi-user CerebroNexus Viewer. In one interactive call it:

1. asks for one or more usernames and hidden login passwords;
2. creates a real encrypted `shinymanager` SQLite database inside the generated
   app;
3. writes the independent database passphrase to an owner-only sibling
   `.auth.env` file; and
4. installs that passphrase in the current R process so the app can run
   immediately.

This workflow requires `shinymanager` (>= 1.1.0), `askpass`, and `openssl`.
Install them in the trusted administration environment that builds the app:

```{r install-auth-dependencies, eval=FALSE, purl=FALSE}
install.packages(c("shinymanager", "askpass", "openssl"))
```

The setup is intentionally interactive. It fails before prompting or creating
output when the dependencies, inputs, destination, or authentication paths are
invalid. Package builds do not run the interactive examples below.

# Interactive quick start

Run this code from a clean working directory in an interactive R or RStudio
session. The example `.crb` is installed with CerebroNexus.

```{r interactive-quick-start, eval=FALSE, purl=FALSE}
example_crb <- system.file(
  "extdata/examples/example.crb",
  package = "CerebroNexus",
  mustWork = TRUE
)
CerebroNexus::createShinyApp(
  cerebro_data = c(example = example_crb),
  result_dir = "my_app",
  auth = TRUE
)
```

The prompts look like this; passwords are collected through `askpass` and are
not echoed:

```text
Username: alice
Password: [hidden]
Confirm password: [hidden]
Add another user? [y/N]: y
Username: bob
Password: [hidden]
Confirm password: [hidden]
Add another user? [y/N]: n
```

Usernames are trimmed and must be unique. A duplicate username is requested
again. Empty or mismatched passwords are requested again for the same user.
Answer `y` or `yes` to add another account; `n`, `no`, or an empty answer
finishes. An empty username or a cancelled password prompt cancels the complete
build instead of publishing a zero-user app.

The account passwords are hashed by `shinymanager`; they are not the database
passphrase. CerebroNexus generates the database passphrase independently from
32 cryptographically random bytes and never prints it.

# Generated files

After a successful quick start, the relevant layout is:

```text
my_app/
├── app.R
├── cerebro_config.rds
└── private-data/
    └── auth/
        └── credentials.sqlite

my_app.auth.env
```

The app config stores only the generated environment-variable name. The
sibling file contains the matching assignment and one final newline:

```text
CEREBRO_AUTH_PASSPHRASE_<16 uppercase hexadecimal characters>=<64 lowercase hexadecimal characters>
```

On POSIX systems, `my_app.auth.env` and `credentials.sqlite` are mode `0600`,
the `auth/` directory is `0700`, and all are owned by the account that built
them. The secret deliberately lives outside `my_app/`: do not move it into the
app, an HTTP resource directory, source control, or a container image.

The SQLite database is encrypted, but it remains writable at runtime because
`shinymanager` records encrypted logs and SQLite may create journals. The
runtime service account therefore needs read/write access to the database and
read/write/search access to its containing directory.

# Run locally

## Current R process

The process that successfully called `createShinyApp()` already has the
generated environment variable, so it can launch the app immediately:

```{r local-run-now, eval=FALSE, purl=FALSE}
shiny::runApp("my_app")
```

## A later R process

A new R or RStudio process must load the durable sibling file first:

```{r local-run-later, eval=FALSE, purl=FALSE}
readRenviron("my_app.auth.env")
shiny::runApp("my_app")
```

`createShinyApp()` does not edit `~/.Renviron`. Loading the deployment-specific
file explicitly avoids leaking its secret to unrelated R sessions.

# Build from your own Cerebro data

`cerebro_data` must be a named vector or list. The names become labels in the
Viewer; each value is a `.crb` or `.rds` path. External H5 or BPCells artifacts
recorded by a `.crb` are copied according to the normal `createShinyApp()`
rules.

```{r own-crb, eval=FALSE, purl=FALSE}
CerebroNexus::createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_seurat.crb"),
  result_dir = "my_app",
  auth = TRUE
)
```

# Copy to another machine

Treat the generated app and secret as a matched deployment pair, but transfer
them separately. This example keeps the secret beside, rather than inside, the
remote app directory. Adapt the host and root to your deployment account:

```{bash copy-to-another-machine, eval=FALSE, purl=FALSE}
set -euo pipefail
ssh deploy@example.org 'install -d -m 0700 "$HOME/cerebronexus"'
rsync -a --delete my_app/ \
  deploy@example.org:cerebronexus/my_app/
rsync -a --chmod=F600 my_app.auth.env \
  deploy@example.org:cerebronexus/my_app.auth.env
```

Load `~/cerebronexus/my_app.auth.env` before starting the remote R process.
Copying only `my_app/` is insufficient: a new process cannot decrypt the
database without the matching assignment. Copying only a different or newly
generated env file is also insufficient because its value will not match the
database.

# Shiny Server with systemd

The following commands illustrate a common Linux installation with the
`shiny` service account, `shiny:shiny` ownership, the `shiny-server` unit, and
apps under `/srv/shiny-server`. Confirm these distribution-specific values
before running the commands.

Stop the service before replacing the deployment. Install the app and secret
to separate locations, then make systemd load the generated assignment before
R starts:

```{bash shiny-server-systemd, eval=FALSE, purl=FALSE}
set -euo pipefail
sudo systemctl stop shiny-server
sudo install -d -m 0755 /srv/shiny-server/my_app
sudo rsync -a --delete my_app/ /srv/shiny-server/my_app/
sudo chown -R shiny:shiny /srv/shiny-server/my_app
sudo chmod 0700 /srv/shiny-server/my_app/private-data/auth
sudo chmod 0600 /srv/shiny-server/my_app/private-data/auth/credentials.sqlite
sudo install -d -m 0700 -o shiny -g shiny /etc/cerebronexus
sudo install -m 0600 -o shiny -g shiny \
  my_app.auth.env /etc/cerebronexus/my_app.auth.env
sudo install -d -m 0755 /etc/systemd/system/shiny-server.service.d
printf '%s\n' \
  '[Service]' \
  'EnvironmentFile=/etc/cerebronexus/my_app.auth.env' |
  sudo tee \
    /etc/systemd/system/shiny-server.service.d/cerebronexus-auth.conf \
    >/dev/null
sudo systemctl daemon-reload
sudo systemctl start shiny-server
sudo systemctl status --no-pager shiny-server
```

The generated assignment uses syntax accepted by both `readRenviron()` and a
systemd `EnvironmentFile`. If local policy prefers an R startup file, the same
owner-only file can instead be selected with an `R_ENVIRON` setting in the
service unit. Do not copy an administrator's personal `~/.Renviron` to the
service account, and do not configure both routes with different values.

After deployment, confirm that the service account can search the app's parent
directories and can read/write `private-data/auth/credentials.sqlite`. Examine
`journalctl -u shiny-server` for the package's redacted authentication error;
do not add commands that print the environment variable itself.

# Docker

Build or use an image that contains `my_app/`, all runtime R packages, and a
Shiny Server or equivalent process listening on port 3838. Exclude the sibling
secret from the build context and every image layer. A `.dockerignore` should
at least contain:

```text
*.auth.env
my_app.auth.env
```

Pass the file only when starting the container. Replace the image name with
your tested image:

```{bash docker-run, eval=FALSE, purl=FALSE}
set -euo pipefail
IMAGE=ghcr.io/your-org/cerebronexus-my-app:latest
docker run --rm --name cerebro-my-app \
  --env-file /secure/my_app.auth.env \
  -p 3838:3838 \
  "$IMAGE"
```

Docker reads the assignment and injects the variable named by
`cerebro_config.rds`; the env file does not need to be mounted inside the
container. Ensure that the container's runtime user can write the bundled
authentication directory and database.

Docker Compose uses the same separately managed file through `env_file`. This
shell chunk creates a valid Compose file and starts it:

```{bash docker-compose, eval=FALSE, purl=FALSE}
set -euo pipefail
cat > compose.yaml <<'YAML'
services:
  cerebro:
    image: ${CEREBRO_IMAGE}
    ports:
      - "3838:3838"
    env_file:
      - /secure/my_app.auth.env
    restart: unless-stopped
YAML
CEREBRO_IMAGE=ghcr.io/your-org/cerebronexus-my-app:latest \
  docker compose up -d
```

Do not replace `env_file` with a literal passphrase in `compose.yaml`. For an
orchestrated production deployment, its native secret manager may inject the
same dynamically named variable instead.

# Rebuild accounts and rotate the database key

## Rebuild while keeping the existing secret

Run `createShinyApp(auth = TRUE)` again with the same `result_dir` to replace
the accounts. If `my_app.auth.env` is still a valid owner-only sibling file,
CerebroNexus reuses its exact variable name and passphrase and builds a new
database containing only the newly entered accounts:

```{r rebuild-accounts, eval=FALSE, purl=FALSE}
CerebroNexus::createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_seurat.crb"),
  result_dir = "my_app",
  auth = TRUE
)
```

The rebuild is staged. Prompt cancellation or another pre-commit failure keeps
the previous app, sibling secret, and process environment. Back up the current
matched app/secret pair before any production account change.

## Intentional key rotation

Rotation creates a new secret and a new database; changing only one makes the
deployment unusable. Use this sequence during a maintenance window:

1. Stop the deployed app.
2. Create a uniquely named protected backup containing both the app directory
   and sibling secret.
3. Move the sibling secret into that backup so its generated path is absent.
4. Run the same `createShinyApp(auth = TRUE)` call and enter the replacement
   accounts. With the final path absent, the call generates a new key and env
   file.
5. Deploy and start the new app and new secret together. If the build fails,
   restore the old secret before restarting the unchanged old app.

```{bash prepare-key-rotation, eval=FALSE, purl=FALSE}
set -euo pipefail
test -d my_app
test ! -L my_app
test -f my_app.auth.env
test ! -L my_app.auth.env
backup_dir="$(mktemp -d './my_app-key-rotation.XXXXXX')"
chmod 0700 "$backup_dir"
cp -a my_app "$backup_dir/my_app"
mv my_app.auth.env "$backup_dir/my_app.auth.env"
printf 'Protected backup: %s\n' "$backup_dir"
```

Keep backups encrypted and access-controlled. Anyone who obtains both the
database and its env file can attempt offline decryption.

# Advanced descriptor mode

Existing deployments may continue to supply an operator-created encrypted
`shinymanager` database. Load its passphrase from a protected file, then pass an
absolute database path and the variable name. This mode validates and copies
the encrypted database but does not create or export a sibling env file:

```{r advanced-descriptor, eval=FALSE, purl=FALSE}
readRenviron("/secure/cerebro-auth.env")
credentials_db <- normalizePath(
  "/secure/credentials.sqlite",
  winslash = "/",
  mustWork = TRUE
)
CerebroNexus::createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_seurat.crb"),
  result_dir = "my_app",
  auth = list(
    provider = "shinymanager",
    credentials = credentials_db,
    passphrase_env = "CEREBRO_AUTH_PASSPHRASE",
    timeout_minutes = 15L
  )
)
```

The named environment variable must already contain at least 32 bytes of
high-entropy secret material and must match the database. `auth = NULL` leaves
the generated Viewer unauthenticated. Explicit `auth = FALSE` is invalid; use
`NULL` when authentication is intentionally disabled.

# Troubleshooting

Inspect the configured variable name without printing its value:

```{r inspect-auth-environment, eval=FALSE, purl=FALSE}
config <- readRDS("my_app/cerebro_config.rds")
env_name <- config$.viewer_auth$passphrase_env
is.na(Sys.getenv(env_name, unset = NA_character_))
```

| Symptom | Likely cause | Safe action |
|---|---|---|
| The app reports a redacted credentials/passphrase error at startup | The configured variable is absent or does not match the database | Stop the app, load the env file paired with this exact app, and restart. Do not print the value while diagnosing it. |
| `readRenviron()` returns `FALSE` or warns | The env file path is missing, unreadable, or malformed | Restore the owner-only sibling file from the matched backup and verify its path. |
| Rebuilding fails before the account prompts | The sibling secret is a symlink, malformed, not owned by the builder, has unsafe POSIX mode, or is under an HTTP resource root | Preserve it for inspection; use a regular owner-owned mode-`0600` file in the frozen parent. Never let the build silently replace it. |
| Authentication works for the builder but not under Shiny Server | The service did not receive the variable or cannot write the SQLite directory/database | Check the unit's `EnvironmentFile`, service account, parent search permissions, auth directory mode/owner, and database mode/owner. Restart after correcting them. |
| A known login password is rejected after a rebuild | Rebuilds replace, rather than merge, the account database | Rebuild again and enter every account that should exist. Login passwords cannot repair a mismatched database passphrase. |
| The sibling env file was lost | The encrypted database key is not recoverable from the database | Restore the matched protected backup, or perform intentional rotation and create a new account database. |

Do not “fix” a wrong key by editing the generated env assignment: the random
value must remain exactly the one used to encrypt the database. A forgotten
user login password and a lost database passphrase are different incidents.

# Security and operational boundaries

The built-in login page is an application access gate, not a complete security
perimeter:

* Terminate TLS/HTTPS at the app or a trusted reverse proxy, and protect every
  untrusted network hop.
* Add rate limiting and request controls at a reverse proxy or WAF; the built-in
  provider does not impose a login-attempt limit by default.
* Use an OIDC/SSO-capable access layer when you require MFA, centralized
  revocation, organization policy, or single sign-on. `shinymanager` does not
  provide those controls.
* Restrict host access. Authentication does not protect files from users who can
  read the deployment account's filesystem.
* Back up the app/database and env file as a matched, encrypted,
  access-controlled pair, and test restoration. Never commit the env file or
  bake it into an image.
* Stop the service or use the documented staged deployment procedure when
  replacing credentials. Do not mix an old app/database with a newly generated
  secret.

# See also

* [Create a self-contained Shiny app from a Cerebro data file](create_a_self_contained_shiny_app.html)
* [Host Cerebro on shinyapps.io](host_cerebro_on_shinyapps.html)
* [Launch Cerebro with a pre-loaded data set](launch_cerebro_with_pre-loaded_data_set.html)
