---
title: "Control access to CerebroNexus with a login page"
author:
  - "Roman Hillje"
  - "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to CerebroNexus with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

Apps generated by `createShinyApp()` are public by default. For a
login-protected Viewer, CerebroNexus needs an encrypted `shinymanager`
credentials database and an environment variable containing the matching
database passphrase.

For new deployments, `provisionViewerAuthentication()` prepares those local
artifacts and returns the strict, secret-free descriptor already accepted by
`createShinyApp()`. It does not configure a server, container, remote host, or
secret store.

The security boundary is easy to remember:

- the generated App contains a copy of the encrypted database;
- `viewer-auth.env` contains the database passphrase and stays outside the App
  tree;
- login passwords exist only in trusted process memory while provisioning;
- Builder and `createShinyApp()` exchange `provision$auth`, not the passwords or
  passphrase; and
- the login page must still be served over HTTPS.

> **Fastest safe path:** prepare a private POSIX parent, call
> `provisionViewerAuthentication()`, load its external env file, pass
> `provision$auth` to `createShinyApp()`, then deploy the App and env file as two
> separately controlled release members.

# One-minute setup

Install the optional providers and load CerebroNexus:

```{r eval=FALSE}
install.packages(c("openssl", "shinymanager"))
library(CerebroNexus)
```

Create a caller-owned private parent. Version 1 provisioning supports POSIX
only and requires the parent to be owned by the current effective user with
mode `0700`.

```bash
install -d -m 0700 /srv/cerebro/private
```

Define the users in trusted R memory. Do not put real passwords in a committed
script, notebook output, log, or command history.

```{r eval=FALSE}
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = c(TRUE, FALSE),
  stringsAsFactors = FALSE
)
```

Provision into a **new** target directory. The target itself must not exist:

```{r eval=FALSE}
provision <- provisionViewerAuthentication(
  accounts = accounts,
  target_dir = "/srv/cerebro/private/viewer-auth",
  timeout_minutes = 15,
  install_env = FALSE
)
rm(accounts)
```

The returned object contains paths and metadata, not login passwords or the
generated passphrase. Load the external handoff file only in the trusted
process that needs it, then build the Viewer:

```{r eval=FALSE}
readRenviron(provision$secret_file)
on.exit(Sys.unsetenv(provision$auth$passphrase_env), add = TRUE)

createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_dataset.crb"),
  result_dir = "/srv/cerebro/apps/viewer",
  auth = provision$auth
)
```

Start the App in the same process, or load the env file again in the runtime
process:

```{r eval=FALSE}
readRenviron("/srv/cerebro/private/viewer-auth/viewer-auth.env")
shiny::runApp("/srv/cerebro/apps/viewer")
```

# What provisioning creates

`target_dir` is a private handoff package. It is sensitive even though one of
its files is encrypted.

| Result or artifact | Purpose | Contains the passphrase? | Goes into the App? |
|---|---|---:|---:|
| `provision$auth` | Strict descriptor for `createShinyApp()` | No | Used at build time |
| `credentials.sqlite` | Encrypted users and authentication state | No plaintext passphrase | Copied into the App |
| `viewer-auth.env` | One environment assignment | **Yes** | **Never** |
| `provision.rds` | Secret-free operation and recovery metadata | No | No |
| `provision$secret_file` | Absolute path to `viewer-auth.env` | The field is only a path | No |
| `provision$manifest_file` | Absolute path to `provision.rds` | No | No |

The result also reports `schema_version`, `user_count`, and whether this call
installed the environment variable in the current process. Printing or saving
the result does not reveal the accounts, login passwords, or passphrase.

```text
accounts in trusted memory
          |
          v
provisionViewerAuthentication()
          |
          +--> encrypted credentials.sqlite
          +--> external viewer-auth.env (secret)
          +--> secret-free provision.rds
          +--> secret-free provision$auth
                         |
                         v
             createShinyApp(auth = provision$auth)
                         |
                         v
       generated App with encrypted database only
```

Keep the complete provisioning target outside the App tree, source control,
web roots, Docker build contexts, ordinary synchronization, and unencrypted
backups.

# How this maps to Builder

The provisioning API and Builder have different responsibilities:

- Builder may collect a one-time `accounts` table and provisioning options in a
  trusted worker.
- Builder calls `provisionViewerAuthentication()` exactly once for the job.
- Builder passes only `provision$auth` to `createShinyApp()`.
- Builder must never persist `accounts`, login passwords, or the generated passphrase.
- Builder may persist the secret-free result fields only when their absolute
  host paths remain meaningful to the same controlled worker/deployment.
- Builder must never copy `viewer-auth.env` into the App, logs, generated code,
  serialized job results, or ordinary artifacts.

A minimal worker flow is:

```{r eval=FALSE}
provision <- provisionViewerAuthentication(
  accounts = submitted_accounts,
  target_dir = private_target,
  passphrase_env = requested_env_name,
  install_env = TRUE
)
submitted_accounts <- NULL

on.exit(Sys.unsetenv(provision$auth$passphrase_env), add = TRUE)
createShinyApp(
  cerebro_data = selected_datasets,
  result_dir = selected_output,
  auth = provision$auth
)
```

Use `install_env = TRUE` only when the same trusted, short-lived worker will
immediately call `createShinyApp()`. The caller owns cleanup of that process
variable after success. For durable services, deploy the external env file or
import its value into the platform secret store; do not depend on the Builder
worker's environment surviving.

# Inputs and preconditions

## Accounts

`accounts` must be a plain data frame with one to 1,000 rows:

- `user`: unique non-empty character values, at most 128 UTF-8 bytes;
- `password`: character values from 12 through 1,024 UTF-8 bytes;
- `admin`: optional non-missing logical values, defaulting to `FALSE`.

Control bytes, missing values, factors, duplicate users, extra columns, and
invalid UTF-8 are rejected. Usernames are trimmed at their edges. Passwords are
passed to the provider without trimming, so generate and transport them
deliberately.

## Target directory

`target_dir` must be a new absolute POSIX path below a pre-existing private
parent. Provisioning verifies the parent owner, exact `0700` mode, path
components, file identities, and absence of the target. It fails closed on
Windows.

The caller must still establish facts that portable R APIs cannot prove, such
as trusted local storage, safe ACLs, absence of a synchronization agent, and no
other uncooperative writer under the same account.

Version 1 creates a new directory only. It does not overwrite an existing
release, rotate a secret in place, or mutate a remote deployment.

## Options

| Argument | Default | Meaning |
|---|---|---|
| `passphrase_env` | `NULL` | Generate an unused `CEREBRO_AUTH_PASSPHRASE_*` name, or request one unused valid name |
| `timeout_minutes` | `15L` | Login inactivity timeout, whole number 1–1440 |
| `install_env` | `FALSE` | Whether to install the new passphrase in this R process after successful publication |

An environment variable already set to the empty string is still occupied and
will not be overwritten.

# Use the result safely

With the default `install_env = FALSE`, explicitly load the handoff file:

```{r eval=FALSE}
readRenviron(provision$secret_file)
stopifnot(nzchar(Sys.getenv(provision$auth$passphrase_env)))
createShinyApp(
  cerebro_data = c(dataset = "dataset.crb"),
  result_dir = "/srv/cerebro/apps/viewer",
  auth = provision$auth
)
Sys.unsetenv(provision$auth$passphrase_env)
```

With `install_env = TRUE`, provisioning installs and verifies the value in the
current process. If a later provisioning step cannot complete, it attempts to
remove only the environment variable owned by that operation.

# Run locally

Every new R process needs the same variable name and value:

```{r eval=FALSE}
options("shinymanager.pwd_failure_limit" = 10L)
readRenviron("/srv/cerebro/private/viewer-auth/viewer-auth.env")
shiny::runApp("/srv/cerebro/apps/viewer")
```

Use a private/incognito browser window to verify that the login page appears,
correct credentials succeed, and a wrong password fails. Test the real HTTPS
deployment URL as well as localhost.

# Deploy with Shiny Server or another R service

Load the file explicitly during trusted service bootstrap:

```{r eval=FALSE}
readRenviron("/absolute/private/viewer-auth.env")
shiny::runApp("/srv/shiny-server/viewer")
```

Deploy only the generated App tree. Keep the environment file outside the App
tree with service-account-only access. The service account needs write access
to the bundled auth directory because `shinymanager` records login state and
SQLite may create journal, WAL, or shared-memory sidecars.

# Deploy with systemd

Add the following to a separately reviewed unit override:

```ini
[Service]
EnvironmentFile=/absolute/private/viewer-auth.env
```

Restart through the operator's normal change process. CerebroNexus does not
edit, reload, or restart the unit.

# Deploy with Docker or Docker Compose

Inject the environment file at runtime rather than copying it into the image:

```bash
docker run --env-file /absolute/private/viewer-auth.env your-viewer-image
```

Equivalently, pass `--env-file /absolute/private/viewer-auth.env` through the
deployment system or configure Compose `env_file`. Do not `COPY` the file into
the image or mount it below the App tree. Keep it outside the Docker build
context and use `.dockerignore` defensively.

# Deploy to a remote host or secret store

Transfer the App and secret handoff through separately controlled encrypted
channels, or import the one-line environment value into the organization's
secret store. Provisioning does not contact another machine or secret service.

Validate the new service before removing the local handoff copy. Retain the old
release secret only for the approved rollback window, then revoke it and its
controlled backups.

# Failure and recovery behavior

Provisioning uses a private lock and staging directory. It publishes the final
target only after the database, secret file, manifest, permissions, identities,
and provider compatibility have been validated.

Most failures remove operation-owned temporary artifacts and return a stable
condition such as `invalid_accounts`, `unsafe_parent`, `database_create_failed`,
or `artifact_publish_failed`. Provider messages and secrets are not included in
the public error text.

If safe cleanup cannot be proven, the condition code is `cleanup_incomplete`
and includes `cause_code` plus an absolute `recovery_path`. Do not blindly
delete that path. Preserve it for operator inspection, confirm ownership and
operation identity, and resolve the underlying filesystem or environment
failure before cleanup.

# Add, remove, or rotate users

Version 1 does not mutate an existing provisioning directory. To change users
or rotate the database passphrase:

1. provision a new target directory with the complete desired account table;
2. build a new App using the new result's `auth` descriptor;
3. deploy the new App and matching external secret as one coordinated release;
4. validate real login, wrong-password rejection, and rollback; and
5. remove the old release only after the approved rollback window.

This approach makes the old release recoverable and avoids half-updated
database/secret pairs.

# Limit failed sign-ins

Set a finite provider failure policy in the App process:

```{r eval=FALSE}
options("shinymanager.pwd_failure_limit" = 10L)
```

Account locking is defense in depth. An attacker can also use it for targeted
denial of service, so configure source-aware rate limiting at the reverse proxy
or ingress and monitor repeated failures. Account for both Shiny HTTP and
WebSocket traffic when tuning the rule.

# Manual database creation (auth-lite compatibility)

The existing auth-lite descriptor also accepts a database created directly
with `shinymanager::create_db()`. Use this path only when provisioning is not
available or an external credential lifecycle already owns the database.

Generate a random passphrase, create a private env file with `Sys.umask("077")`,
load it with `readRenviron()`, and call:

```{r eval=FALSE}
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = c(TRUE, FALSE),
  stringsAsFactors = FALSE
)

shinymanager::create_db(
  credentials_data = accounts,
  sqlite_path = "/srv/cerebro/private/credentials.sqlite",
  passphrase = Sys.getenv("CEREBRO_AUTH_PASSPHRASE")
)
rm(accounts)

createShinyApp(
  cerebro_data = c(dataset = "dataset.crb"),
  result_dir = "/srv/cerebro/apps/viewer",
  auth = list(
    credentials = "/srv/cerebro/private/credentials.sqlite",
    passphrase_env = "CEREBRO_AUTH_PASSPHRASE",
    timeout_minutes = 15
  )
)
```

For example, OpenSSL can generate 32 random bytes:

```bash
openssl rand -base64 32
```

# Troubleshooting

| Symptom | Likely cause | Action |
|---|---|---|
| `unsafe_parent` | Parent ownership, mode, identity, path style, or target state is unsafe | Use a new target below a trusted current-user-owned `0700` POSIX parent |
| `target_exists` | Provisioning never overwrites a target | Choose a new versioned target directory |
| `missing_dependency` | `openssl` or `shinymanager >= 1.1.0` is unavailable | Install the optional providers in the worker environment |
| `environment_conflict` | Requested variable already exists, even as an empty value | Choose an unused name or unset the old variable deliberately |
| `database_create_failed` | Provider rejected the accounts or could not create the DB | Recheck the account table and private filesystem; public errors intentionally hide provider output |
| `cleanup_incomplete` | Cleanup ownership or a filesystem/environment action could not be proven | Preserve and inspect `recovery_path`; do not recursively delete it blindly |
| Build cannot decrypt the database | Runtime variable is absent or belongs to a different release | Load the exact env file paired with the selected provisioning target |
| App contains no login page | `auth = NULL` or the wrong App tree was deployed | Build and deploy using `provision$auth` |
| App cannot record login state | Service account cannot write the bundled auth directory | Correct narrowly scoped runtime ownership and permissions |

# Security checklist

Before release, verify that:

- provisioning ran on a trusted POSIX host below a private parent;
- the App contains no `viewer-auth.env`, passphrase, or plaintext login password;
- `viewer-auth.env` remains outside the App tree and ordinary publication paths;
- Builder retained only the secret-free descriptor/result fields it needs;
- the runtime received the environment value paired with the bundled database;
- correct login succeeds and incorrect login fails over HTTPS;
- runtime permissions allow auth-state writes without exposing unrelated paths;
- rate limiting and monitoring complement account locking; and
- rotation, backup retention, and the rollback window are defined.

# See also

- [Create a self-contained Shiny App](create_a_self_contained_shiny_app.html)
- [Host CerebroNexus on shinyapps.io](host_cerebro_on_shinyapps.html)
- [`provisionViewerAuthentication()`](../reference/provisionViewerAuthentication.html)
- [`shinymanager::create_db()`](https://datastorm-open.github.io/shinymanager/reference/create_db.html)
