---
title: "Control access to CerebroNexus with a login page"
author:
  - "Roman Hillje"
  - "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
toc-title: "Contents"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to CerebroNexus with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

<style>
.main-container {
  width: auto !important;
  max-width: none !important;
  margin: 0;
  padding: 0 clamp(1rem, 4vw, 4rem);
}
pre {
  overflow-x: auto;
}
img {
  display: block;
  width: 100%;
  height: auto;
  margin: 1.5rem auto;
}
</style>

# Overview

`provisionViewerAuthentication()` turns an in-memory account table into:

- an encrypted `shinymanager` database;
- an external passphrase file;
- secret-free recovery metadata; and
- `provision$auth`, ready for `createShinyApp()`.

The App receives only the encrypted database. Login passwords and the
passphrase stay outside the App tree. HTTPS is still required.

![Provisioning validates, stages, publishes, and keeps recovery evidence when
safe cleanup cannot be proved.](img/auth-provisioning-validation.svg)

# One-minute setup

**Run this as:** the release operator or a trusted, short-lived build worker.

**Run it on:** one trusted POSIX build host with R, CerebroNexus, the source CRB
files, and a current-user-owned private filesystem area.

**What happens:** provisioning creates a versioned authentication handoff;
`createShinyApp()` then copies only its encrypted database into the App.

**Verify:** inspect the returned paths, build the App, and test one valid and one
invalid login before deployment.

Install the optional providers:

```{r eval=FALSE}
# R console — run this once to install and load the provisioning tools.
install.packages(c("openssl", "shinymanager"))
library(CerebroNexus)
```

Create a private POSIX parent:

```bash
# Terminal — run this on the trusted POSIX build host.
install -d -m 0700 /srv/cerebro/private
```

Define the users in trusted memory:

```{r eval=FALSE}
# R console — enter this only in the trusted provisioning session.
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = c(TRUE, FALSE),
  stringsAsFactors = FALSE
)
```

Provision into a new target directory:

```{r eval=FALSE}
# R console — run this to create a new authentication handoff directory.
provision <- provisionViewerAuthentication(
  accounts = accounts,
  target_dir = "/srv/cerebro/private/viewer-auth",
  timeout_minutes = 15,
  install_env = FALSE
)
rm(accounts)
```

Load the secret and build the Viewer:

```{r eval=FALSE}
# R console — run this in the same trusted process that builds the App.
readRenviron(provision$secret_file)
on.exit(Sys.unsetenv(provision$auth$passphrase_env), add = TRUE)

createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_dataset.crb"),
  result_dir = "/srv/cerebro/apps/viewer",
  auth = provision$auth
)
```

Start the App with the same release secret:

```{r eval=FALSE}
# R console — run this in the App process to start the local Viewer.
readRenviron("/srv/cerebro/private/viewer-auth/viewer-auth.env")
shiny::runApp("/srv/cerebro/apps/viewer")
```

# What provisioning creates

| Result | Purpose | Secret? | In the App? |
|---|---|---:|---:|
| `provision$auth` | Build descriptor | No | Used at build time |
| `credentials.sqlite` | Encrypted users and login state | No plaintext | Copied |
| `viewer-auth.env` | Database passphrase | **Yes** | **Never** |
| `provision.rds` | Ownership and recovery metadata | No | No |

Treat the complete target directory as sensitive: it contains the database and
its key. Keep it outside Git, web roots, Docker contexts, ordinary sync, and
unencrypted backups.

# Inputs and preconditions

## Accounts

`accounts` is a plain data frame with 1–1,000 rows:

| Column | Rule |
|---|---|
| `user` | Unique, non-empty, at most 128 UTF-8 bytes |
| `password` | 12–1,024 UTF-8 bytes; not trimmed |
| `admin` | Optional logical; defaults to `FALSE` |

Missing values, factors, extra columns, duplicate users, control bytes, and
invalid UTF-8 fail before any files are created.

## Target directory

`target_dir` must be a new absolute POSIX path below an existing, current-user
owned `0700` parent. Windows fails closed in version 1.

The caller must separately ensure the parent is on trusted local storage, has
safe ACLs, and is not managed by an untrusted writer or synchronization agent.

## Options

| Argument | Default | Meaning |
|---|---|---|
| `passphrase_env` | `NULL` | Generate an unused variable name |
| `timeout_minutes` | `15L` | Whole number from 1 to 1440 |
| `install_env` | `FALSE` | Install the secret in this R process |

Even an empty existing environment variable is treated as occupied.

# Use the result safely

**Run this as:** the process that immediately builds the App.

**Run it on:** the same trusted host where the provisioning paths are valid.

**What happens:** the env file loads the generated passphrase into this process;
the descriptor points `createShinyApp()` to the encrypted database.

**Verify:** confirm the variable is non-empty, complete the build, and unset the
variable or terminate the dedicated worker.

With the default `install_env = FALSE`:

```{r eval=FALSE}
# R console — run this in the trusted process that consumes the result.
readRenviron(provision$secret_file)
stopifnot(nzchar(Sys.getenv(provision$auth$passphrase_env)))
createShinyApp(
  cerebro_data = c(dataset = "dataset.crb"),
  result_dir = "/srv/cerebro/apps/viewer",
  auth = provision$auth
)
Sys.unsetenv(provision$auth$passphrase_env)
```

The returned object and `provision.rds` contain paths and counts, never the
account table, login passwords, or passphrase.

# Run locally

**Run this as:** the local App operator.

**Run it on:** a machine holding the generated App and the matching provisioning
release.

**What happens:** the R process loads the release secret, unlocks the bundled
database, and exposes the login page on the local Shiny port.

**Verify:** open <http://localhost:3838>, confirm valid login succeeds and invalid
login fails, then stop the R process with Ctrl+C.

```{r eval=FALSE}
# R console — run this in every local App process.
options("shinymanager.pwd_failure_limit" = 10L)
readRenviron("/srv/cerebro/private/viewer-auth/viewer-auth.env")
shiny::runApp("/srv/cerebro/apps/viewer")
```

Test a valid and invalid login in a private browser window, then repeat on the
real HTTPS URL.

![The App and external passphrase remain separate until the trusted runtime
loads both.](img/auth-deployment-boundary.svg)

# Deploy with Shiny Server or another R service

**Run this as:** the service operator who controls the App startup script.

**Run it on:** the runtime host, using the actual service account and its
protected env-file path.

**What happens:** the startup process loads the passphrase before running the
deployed App; the env file itself remains outside the App tree.

**Verify:** inspect the service environment and logs, then test the real HTTPS
URL with both a valid and invalid password.

```{r eval=FALSE}
# R startup file — save this in the reviewed service startup script.
readRenviron("/absolute/private/viewer-auth.env")
shiny::runApp("/srv/shiny-server/viewer")
```

Deploy only the App tree. Grant the runtime account write access only to the
bundled auth directory so SQLite can record login state.

# Deploy with systemd

**Run this as:** the systemd service administrator.

**Run it on:** the host running the Shiny service.

**What happens:** systemd reads `EnvironmentFile=` before launching the service,
so every App process inherits the required variable.

**Verify:** restart the service, inspect status and logs, then test the deployed
login page.

```ini
# systemd override — save this in the editor opened by `sudo systemctl edit shiny-server`.
[Service]
EnvironmentFile=/absolute/private/viewer-auth.env
```

CerebroNexus does not edit or restart the service. The operator applies the
change explicitly:

```bash
# Terminal — run these commands after saving the systemd override.
sudo systemctl daemon-reload
sudo systemctl restart shiny-server
sudo systemctl status shiny-server
sudo journalctl -u shiny-server -f
```

# Deploy with Docker or Docker Compose

**Run this as:** a user allowed to run Docker.

**Run it on:** the container host that holds the image and external env file.

**What happens:** Docker reads the host env file, injects its assignment into the
container, and publishes container port 3838 without adding the secret to the
image.

**Verify:** open <http://localhost:3838>, test both login outcomes, and follow the
container logs.

```bash
# Terminal — run these commands on the Docker host.
docker run --name cerebro --detach --publish 3838:3838 \
  --env-file /absolute/private/viewer-auth.env \
  your-viewer-image

# Follow startup and authentication logs.
docker logs -f cerebro

# Stop the container when this release is no longer serving traffic.
docker stop cerebro
```

Compose can use the same file through `env_file`. Keep it outside the build
context and the App tree.

# Deploy to a remote host or secret store

**Run this as:** the destination operator or the deployment platform's release
identity.

**Run it on:** the destination host or secret-management control plane.

**What happens:** the App and secret travel through separate controlled
channels; the destination runtime reconstructs the named environment variable.

**Verify:** confirm the platform reports the expected variable name, test the
new HTTPS login page, and retain the previous pair only for the rollback window.

Transfer the App and secret through separate encrypted channels, or import the
single environment value into the platform secret store. On another machine,
load that value before starting the App. Validate the new service before
deleting the handoff copy.

# Failure and recovery behavior

Provisioning publishes only after the database, secret, manifest, permissions,
identities, and provider compatibility pass.

Most failures remove only operation-owned temporary files. If cleanup cannot
be proved, the error is `cleanup_incomplete` and includes `cause_code` plus a
`recovery_path`. Preserve that path for inspection; never delete it blindly.

# Add, remove, or rotate users

**Run this as:** the credential administrator or a trusted automation process.

**Run it on:** the trusted build host, using a new target directory.

**What happens:** provisioning creates a complete replacement release; it never
edits the active database or passphrase in place.

**Verify:** deploy the new App and matching secret together, test both login
outcomes, and prove rollback before retiring the old pair.

Provision a new directory, build a new App, and deploy the App and matching
secret together. Keep the old pair only for the approved rollback window.

# Limit failed sign-ins

```{r eval=FALSE}
# R console — place this in the real App startup process or its `.Rprofile`.
options("shinymanager.pwd_failure_limit" = 10L)
```

Add source-aware rate limiting and monitoring at the reverse proxy; account
locking alone can be abused for denial of service.

# Manual database creation (auth-lite compatibility)

**Run this as:** the credential administrator when provisioning is unavailable
or another approved lifecycle already owns the database.

**Run it on:** a trusted host with a private source-database directory.

**What happens:** you manually reproduce the auth-lite inputs that provisioning
normally creates: an env file, encrypted database, and strict descriptor.

**Verify:** build the App with that descriptor and test valid and invalid login
using the matching env file.

Use this path only when an external lifecycle already owns the database:

```{r eval=FALSE}
# R console — run this to create the manual external env file.
old_umask <- Sys.umask("077")
on.exit(Sys.umask(old_umask), add = TRUE)
writeLines(
  "CEREBRO_AUTH_PASSPHRASE=replace-with-a-random-value",
  "/srv/cerebro/private/viewer-auth.env"
)
Sys.chmod("/srv/cerebro/private/viewer-auth.env", "0600")
```

```{r eval=FALSE}
# R console — run this to create the database and build the App manually.
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = c(TRUE, FALSE),
  stringsAsFactors = FALSE
)
shinymanager::create_db(
  credentials_data = accounts,
  sqlite_path = "/srv/cerebro/private/credentials.sqlite",
  passphrase = Sys.getenv("CEREBRO_AUTH_PASSPHRASE")
)
rm(accounts)

createShinyApp(
  cerebro_data = c(dataset = "dataset.crb"),
  result_dir = "/srv/cerebro/apps/viewer",
  auth = list(
    credentials = "/srv/cerebro/private/credentials.sqlite",
    passphrase_env = "CEREBRO_AUTH_PASSPHRASE",
    timeout_minutes = 15
  )
)
```

```bash
# Terminal — run this command and save the result privately.
openssl rand -base64 32
```

# Troubleshooting

| Symptom or code | Action |
|---|---|
| `unsafe_parent` | Use a trusted current-user-owned `0700` parent |
| `target_exists` | Choose a new versioned target |
| `missing_dependency` | Install `openssl` and `shinymanager >= 1.1.0` |
| `environment_conflict` | Choose an unused variable name |
| `cleanup_incomplete` | Preserve and inspect `recovery_path` |
| Build cannot decrypt | Load the env file paired with that target |
| App has no login page | Build and deploy with `provision$auth` |
| SQLite cannot write | Fix access only for the bundled auth directory |

# Security checklist

- Provision on a trusted POSIX host.
- Keep passwords, passphrase, and env file outside the App and Git.
- Persist only secret-free result fields; never persist the account table or passphrase.
- Pair each App release with exactly one matching secret.
- Verify valid login, invalid login, HTTPS, write access, and rollback.
- Add monitoring and rate limiting.

# See also

- [Create a self-contained Shiny App](create_a_self_contained_shiny_app.html)
- [Host CerebroNexus on shinyapps.io](host_cerebro_on_shinyapps.html)
- [`provisionViewerAuthentication()`](../reference/provisionViewerAuthentication.html)
- [`shinymanager::create_db()`](https://datastorm-open.github.io/shinymanager/reference/create_db.html)
