---
title: "Create a self-contained Shiny app from a Cerebro data file"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Create a self-contained Shiny app from a Cerebro data file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`createShinyApp()` bundles a Cerebro v1.4 Shiny app into a single output directory. It copies the Shiny sources shipped under `inst/shiny/v1.4/`, the requested `.crb` (or `.rds`) data file(s), and the `inst/extdata/` reference files, and writes an `app.R` that wires everything together with a pre-built `Cerebro.options` list. The result is a directory you can serve directly with `shiny-server`, drop behind `rsconnect`/Posit Connect, or launch locally with `shiny::runApp()` — no further calls into `CerebroNexus` are required at runtime.

This is the recommended path when you want to hand a colleague a runnable copy of Cerebro pre-loaded with a specific data set without making them install the package themselves, pin the Shiny sources at a known revision alongside the data, or deploy to a host that doesn't run R interactively (`shinyapps.io`, Docker images, etc.).

# Setup

The package ships an example `.crb` in `inst/extdata/v1.4/`, which we use throughout this vignette.

```{r}
library(CerebroNexus)

crb <- system.file("extdata/v1.4/example.crb", package = "CerebroNexus")
file.exists(crb)
#> [1] TRUE
```

# Quick start

`cerebro_data` must be a **named** character vector (or list). Names must be non-missing and unique because they become the dataset labels shown in the Load Data tab; the values are paths to `.crb` (or `.rds`) files.

```{r}
out_dir <- file.path(tempdir(), "cerebro_app")

createShinyApp(
  cerebro_data = c("PBMC example" = crb),
  result_dir   = out_dir,
  launch_browser = FALSE
)
```

After this call, `out_dir` contains the generated `app.R`, a copy of the Shiny sources under `shiny/`, the `inst/extdata/` reference files, the data file(s) under `data/`, and a `cerebro_config.rds` holding the serialized `Cerebro.options`.

Launch it locally:

```{r, purl=FALSE}
shiny::runApp(out_dir)
```

# Common parameters

| argument | purpose |
|---|---|
| `cerebro_data` | named vector/list of `.crb` (or `.rds`) paths |
| `result_dir` | output directory (required) |
| `overwrite` | replace `result_dir` after a complete staged build; with `FALSE`, the destination must be absent or empty |
| `max_request_size` | upload size cap in MB; defaults to `8000` |
| `port`, `host` | binding for the generated `app.R`; defaults to `8080` / `127.0.0.1` |
| `launch_browser`, `quiet`, `display_mode` | forwarded to `shiny::runApp()` in `app.R` |
| `welcome_message` | text shown in the Load Data tab |
| `colors` | optional named list of palettes (one entry per dataset name) |
| `cerebro_options` | extra entries merged into `Cerebro.options` |
| `crb_pick_smallest_file`, `show_upload_ui` | forwarded into `Cerebro.options` |
| `point_size`, `variable_to_compare` | forwarded into `Cerebro.options` |

# Bundling multiple datasets （available in V2.0.0）

The names of `cerebro_data` are what the user picks from inside Cerebro. You can also supply matching `colors` so each dataset gets a deterministic palette.

```{r}
crb_pbmc <- system.file("extdata/v1.4/example.crb", package = "CerebroNexus")
# crb_other <- "/path/to/another_dataset.crb"

createShinyApp(
  cerebro_data = c(
    "PBMC example"   = crb_pbmc
    # , "My dataset" = crb_other
  ),
  result_dir = file.path(tempdir(), "cerebro_app_multi"),
  colors = list(
    "PBMC example" = list(sample = c(pbmc_1 = "#1f77b4"))
  ),
  welcome_message = "Welcome to my Cerebro deployment.",
  launch_browser = FALSE
)
```

# Sibling files: `.bpcells/` and `.h5`

If your `.crb` was exported with an external expression backend
(`expression_matrix_mode = "bpcells"` or `"h5"` in
`exportFromSeurat()`), its descriptor records the matrix location relative to
the `.crb`. Fresh exports normally use a sibling `<stem>.bpcells/` directory or
`<stem>.h5` file, but the descriptor remains authoritative if the `.crb` is
renamed.

`createShinyApp()` copies the tagged file or directory to the same relative
location under `result_dir/data/`. The location must be a portable relative
path; missing backends and conflicting bundle targets are errors rather than
silent overwrites. A corresponding global runtime override retains its existing
precedence; when configured, its existence is not checked at build time, the
sidecar is not copied, and the deployed app depends on that host path. Because
the override is global, it may serve only one effective CRB consumer in a
multi-dataset app. Missing optional spatial images are omitted with a warning.

Before copying, `createShinyApp()` resolves every CRB descriptor and sidecar,
and plans every existing spatial image and destination path together. It
rejects exact and parent/child target collisions as well as backend paths that
resolve through symbolic links. The complete app is assembled in a private
sibling directory and published only after all writes succeed. Consequently, a
failed rebuild leaves the previous deployment untouched. `overwrite = FALSE`
is intentionally conservative: a non-empty destination is rejected before the
build starts rather than merged with a partially known old app.

On POSIX systems, the staging directory is mode `0700` while data is copied.
Replacing an existing app retains its root permission bits; platform-specific
ACLs, ownership changes and security labels remain the deployment system's
responsibility. Inputs are validated and then copied without immutable
snapshots, so do not modify a CRB or backend while `createShinyApp()` is running.

# Forwarding extra Cerebro options

Anything you want to surface in `Cerebro.options` that isn't already a top-level argument can go through `cerebro_options`.

```{r}
createShinyApp(
  cerebro_data = c("PBMC example" = crb),
  result_dir   = file.path(tempdir(), "cerebro_app_opts"),
  cerebro_options = list(
    exclude_trivial_metadata = TRUE
  ),
  launch_browser = FALSE
)
```

`mode`, `cerebro_version`, `crb_file_to_load`, and `cerebro_root` are filled in by the function itself — anything you supply for those keys will be overwritten.

# Deploying the bundle

The output directory is self-contained: shipping the folder is all that's required. Three common targets:

```{r, purl=FALSE}
# 1. Local
shiny::runApp(out_dir)

# 2. shiny-server / Posit Connect
# Drop the directory under the server's app root (e.g. /srv/shiny-server/),
# or use the Posit Connect publishing UI pointing at `app.R`.

# 3. shinyapps.io
rsconnect::deployApp(appDir = out_dir)
```

# Reference

- `?createShinyApp`
- `vignette("cerebroApp_workflow_Seurat")` for end-to-end Seurat → `.crb` → app
- `vignette("host_cerebro_on_shinyapps")` for shinyapps.io specifics
