---
title: "Demo dataset: Omnibus (all modalities)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: Omnibus (all modalities)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

|  |  |
|---|---|
| **Technology** | **Fully synthetic** — fabricated in R, no real assay |
| **Organism / tissue** | hg (declared) / a made-up inflamed-tissue section with infiltrating lymphocytes |
| **Source** | None. 100% fabricated from `set.seed(20260723)`. |
| **License** | None required — freely redistributable. |
| **Used in** | *Every* tab: Projection, Groups, Marker genes, Immune repertoire, HLA & TCR Motifs, Spatial, Trekker, Linked views. |
| **How to open it** | It ships with CerebroNexus (~0.8 MB) — launch the app and pick **Omnibus (all modalities)** from the data-set selector. Nothing to download or build. |

Unlike every other demo, this one is **not real data**. It exists for two reasons:

1. **Coverage** — a single `.crb` that carries *every* modality the app supports, so you can exercise the whole application from one object instead of switching data sets.
2. **Coexistence regression test** — it is the only demo carrying **both** a standard `spatial` slot **and** a `trekker` slot, so it guards that the two coexist as their own tiled panels in **Linked views** (see [Spatial and Trekker in one view](#spatial-and-trekker-in-one-view)).

Because the data is fabricated, **no number here is a biological claim**: the HLA
genotypes, dextramer calls, CDR3 sequences, Moran's I values, and QC metrics
exist only to drive the app's code paths. The object says so in its own
`technical_info$tcr_selection_detail`.

# What one object carries

| Modality | How it is fabricated | Tab it lights up |
|---|---|---|
| Expression + UMAP / tSNE / PCA | marker-driven Poisson counts (sparse), per-cluster Gaussian embeddings | Projection, Gene expression |
| Groups | `sample`, `cell_type`, `cluster`, `region`, `dextramer_*`, `restriction_in_genotype` | Groups, Overview |
| Marker genes | per-cell-type tables (real symbols + synthetic) | Marker genes |
| TCR immune repertoire | 6 **convergent** CDR3 families (Hamming-1 variants) on the T compartment | Immune repertoire, Linked views (clone axis) |
| HLA typing | three synthetic donor genotypes (one homozygous, one half-called locus) | HLA & TCR Motifs |
| Spatial (standard) | three tissue sections, **each with an embedded synthetic H&E** | Spatial, Linked views (spatial panel) |
| Trekker | a positioned single-cell subset + confidence + per-cell fields + evidence + Moran's I + QC | Trekker |

# Spatial and Trekker in one view

This is the only demo carrying **every** space at once — UMAP, the standard
spatial map, the Trekker physical map, and the clonal axis — so **Linked views**
gives each its **own panel** and tiles them to fill the screen: two spaces sit
side by side, four fall into a 2×2, and three take a rotated-"品" (UMAP centred on
the left, the other two stacked on the right). Every panel is on screen at once,
and the linked selection stays keyed on the same cells across all of them, so a
lasso in the UMAP lights up the same cells in tissue, in the Trekker map, and on
the clonal axis simultaneously. Colouring by a Trekker field (spatial purity,
cross-space concordance, position confidence) works alongside the spatial image.

That coexistence is the whole point: because no *real* demo carries both
modalities at once, this fabricated one is what keeps them from colliding as the
app changes. (It briefly did collide — until 2026-07-23 a standard spatial map
silently hid the Trekker one in Linked views; the fix gave Trekker its own panel,
so both now appear. The dedicated **Spatial** and **Trekker** tabs were never
affected.)

# Rebuild it yourself (optional)

You don't need to — the `.crb` already ships with the package. But because it is
100% synthetic, you can regenerate it from scratch with no downloads. From a clone
of the CerebroNexus repository:

```bash
Rscript data-raw/build_omnibus_demo.R
```

The script synthesises every modality deterministically, runs a verification gate
(shape, sparsity, alignment, three-state restriction column, complete Trekker QC),
and writes the `.crb` back into the package's own data directory so the rebuilt
demo replaces the shipped one. (That repo path is where the *package* keeps it; if
you just want to use the file, see the next section.)

# How each modality is fabricated

The script is linear and commented; the essentials:

**Cells & clusters.** 3,000 cells across 3 synthetic donors, assigned to 7 cell
types by probability, then to 12 finer clusters (kept 0-based for Trekker).

**Expression.** A gene panel of 175 markers (5 real symbols + synthetic per type)
plus 400 background genes. A lambda matrix elevates each cell's own-type markers;
`rpois()` draws counts; library-size-normalised `log1p` is the "data" layer.
Zeros stay zeros, so the block stays a sparse `dgCMatrix`.

**Projections.** UMAP and tSNE are per-cluster Gaussian blobs (consistent with the
marker structure by construction); PCA is `prcomp` on the expression block.

**TCR.** Six germline β/α CDR3s each seed a **convergent family** of Hamming-1
variants — the shape an antigen-selected repertoire has and an unselected one
lacks, which is why the motif network is legible. Clone sizes follow a power law
(a few large clones, a long singleton tail).

**HLA.** Three donor genotypes, chosen so one donor is homozygous `A*03:01` and
one has a half-called HLA-B locus — this is what exercises the three-state
`restriction_in_genotype` column (`yes` / `no` / `unknown`), the same honesty
logic the real dextramer demo uses.

**Spatial.** Each donor's cells are laid out in a 2-D tissue by region band, and a
small synthetic H&E (drawn with `grDevices::png`, eosin blush + haematoxylin
nuclei) is embedded as a base64 `data:` URI with its coordinate-space bounds.
Three donors → three histology images, all self-contained.

**Trekker.** A ~70% positioned subset carries physical `x`/`y`, UMAP coordinates,
0-based clusters, three per-cell fields (`mk_field()` → 0–255 vectors), per-nucleus
confidence, a handful of synthetic positioning-evidence JPEGs, Moran's I rows, and
a complete vendor-style QC block (the Trekker page and the Linked-views info modal
call `.toFixed()` on those QC fields, so all of them must be present).

# Open it in the app

The demo ships with CerebroNexus, so the simplest path is: launch the app and
choose **Omnibus (all modalities)** from the data-set selector — no file handling
at all.

If instead you have a `.crb` sitting in your working directory (e.g. one you
rebuilt above, or your own data set) and want to wrap it into a standalone app,
point `createShinyApp()` at that local file:

```r
library(CerebroNexus)
createShinyApp(
  # a plain path to wherever YOUR .crb lives — here, the working directory
  cerebro_data = c("Omnibus (all modalities)" = "demo_omnibus.crb"),
  result_dir = "cerebro_app"
)
shiny::runApp("cerebro_app")
```

# Full build code

```{r code=xfun::read_utf8('../data-raw/build_omnibus_demo.R')}
```
