---
title: "Demo dataset: Mouse hippocampus (Slide-seq v2)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: Mouse hippocampus (Slide-seq v2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> One of four real spatial-transcriptomics demos, each on a different platform.
> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

| | |
|---|---|
| **Technology** | Slide-seq v2 (bead, 10 µm) |
| **Organism / tissue** | mouse / hippocampus |
| **Source** | Stickels et al. 2021 · [doi:10.1038/s41587-020-0739-1](https://doi.org/10.1038/s41587-020-0739-1), via `ssHippo.SeuratData` |
| **License** | SeuratData terms; original data CC-BY (Broad / Macosko lab) |
| **Used in** | [Spatial](spatial_analysis.html) tab |
| **Output** | shipped as `inst/extdata/v1.4/demo_spatial_slideseq.crb` (~1.6 MB) |

This is the end-to-end proof that `exportFromSeurat` handles **non-v5** (v4-era)
spatial objects: `ssHippo` ships as a Seurat v3.1.4 object, so
`UpdateSeuratObject()` repairs it before anything else runs — an intentionally
different code path from the other three (Seurat-5-native) spatial demos.

# Download raw data

```r
SeuratData::InstallData("ssHippo")
data("ssHippo", package = "ssHippo.SeuratData")
obj <- get("ssHippo")
```

# Process in Seurat

```r
obj <- SeuratObject::UpdateSeuratObject(obj)   # v3.1.4 -> current SlideSeq class
assay <- "Spatial"

set.seed(42)
keep <- sample(colnames(obj), min(5000, ncol(obj)))   # down-sample first: 53k beads
obj <- subset(obj, cells = keep)

obj <- NormalizeData(obj, assay = assay) |>
  FindVariableFeatures(assay = assay, nfeatures = 2000) |>
  ScaleData(assay = assay) |>
  RunPCA(assay = assay, npcs = 30) |>
  RunUMAP(dims = 1:20) |>
  FindNeighbors(dims = 1:20) |>
  FindClusters(resolution = 0.5)
obj$cluster <- factor(paste0("C", as.integer(obj$seurat_clusters)))
```

- **sampling**: 5,000 beads sampled from ~53,173, *before* processing (the source
  is unclustered at that scale).
- **cell-type field**: `cluster` (source ships no cell-type labels).
- **embedded image**: **none — and this is correct, not an omission.** The
  `SlideSeq` S4 class carries only bead `coordinates`, no tissue raster; the
  bead scatter is the complete spatial view.

# Export to Cerebro (`.crb`)

```r
library(CerebroNexus)
exportFromSeurat(
  object = obj,
  assay = "Spatial",
  slot = "data",
  file = "slideseq_mouse_hippocampus.crb",
  experiment_name = "Slide-seq v2 mouse hippocampus",
  organism = "mm",
  groups = c("cluster"),
  nUMI = "nUMI",
  nGene = "nGene"
)
```

# Build & launch the Shiny app

```r
library(CerebroNexus)

crb <- "slideseq_mouse_hippocampus.crb"

createShinyApp(
  cerebro_data = c("Mouse hippocampus (Slide-seq v2)" = crb),
  result_dir   = "my_cerebro_app",
  launch_browser = FALSE
)

# Local:
shiny::runApp("my_cerebro_app")

# shinyapps.io:
# rsconnect::deployApp(appDir = "my_cerebro_app")
```

Open the **Spatial** tab and pick "Mouse hippocampus (Slide-seq v2)" from the
dataset switcher.

# Full build code

Produced by `build_slideseq()`, one of four functions in
[`data-raw/build_spatial_demos.R`](https://github.com/duocang/CerebroNexus/blob/master/data-raw/build_spatial_demos.R)
(shared download/sampling helpers are defined earlier in that file):

```{r code=local({src <- xfun::read_utf8("../data-raw/build_spatial_demos.R"); s <- grep("^## 2\\. Slide-seq", src); e <- grep("^## 3\\. MERFISH", src)[1] - 2; src[s:e]})}
```
