---
title: "Demo dataset: Mouse brain (10x Xenium)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: Mouse brain (10x Xenium)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> One of four real spatial-transcriptomics demos, each on a different platform.
> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

| | |
|---|---|
| **Technology** | 10x Xenium (in-situ single-cell imaging) |
| **Organism / tissue** | mouse / brain (cortex + hippocampus) |
| **Source** | 10x Genomics public [`Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP`](https://www.10xgenomics.com/datasets/fresh-frozen-mouse-brain-replicates-1-standard) |
| **License** | 10x Genomics public dataset terms |
| **Used in** | [Spatial](spatial_analysis.html) tab · [Linked views](../) |
| **Output** | shipped as `inst/extdata/v1.4/demo_spatial_xenium.crb` (~3.4 MB) |

A real cortex + hippocampus sub-section (248-gene Mouse Brain panel) with
recognisable brain structure — not a rectangular field-of-view crop.

# Download raw data

The build script auto-fetches this (`ensure_unzipped()`); the manual equivalent:

```bash
mkdir -p data-raw/xenium
BASE=https://cf.10xgenomics.com/samples/xenium/1.0.2/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP
curl -fL -o data-raw/xenium/brain.zip \
  "$BASE/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP_outs.zip"
unzip -oq data-raw/xenium/brain.zip -d data-raw/xenium/brain
```

# Process in Seurat

Built from the raw 10x outs bundle
(`cell_feature_matrix.h5` + `cells.csv.gz` + `morphology_focus.ome.tif`)
directly, **not** `LoadXenium()` — the public bundle ships the `.h5` matrix,
not the mtx directory `LoadXenium` expects:

```r
set.seed(42)
mat <- Read10X_h5("data-raw/xenium/brain/cell_feature_matrix.h5")
cells <- read.csv(gzfile("data-raw/xenium/brain/cells.csv.gz"))
rownames(cells) <- cells$cell_id
common <- intersect(colnames(mat), cells$cell_id)

obj <- CreateSeuratObject(counts = mat[, common], assay = "Spatial")
xy <- data.frame(x = cells[common, "x_centroid"], y = cells[common, "y_centroid"],
                  row.names = common)

obj <- NormalizeData(obj) |> FindVariableFeatures() |> ScaleData() |>
  RunPCA(npcs = 30) |> RunUMAP(dims = 1:20) |>
  FindNeighbors(dims = 1:20) |> FindClusters(resolution = 0.5)
obj$cluster <- factor(paste0("C", as.integer(obj$seurat_clusters)))
# stratified down-sample to <= 5,000 by obj$cluster

cents <- CreateCentroids(as.matrix(xy[colnames(obj), c("x", "y")]))
obj[["fov"]] <- CreateFOV(list(centroids = cents), type = "centroids", assay = "Spatial")
```

- **sampling**: ≤ 5,000 cells stratified by cluster (here 5,000 of 36,602);
  248-gene panel kept whole.
- **cell-type field**: `cluster` (source ships no cell-type labels).
- **embedded image**: **real DAPI morphology** — first channel of the OME-TIFF,
  contrast-stretched, mapped into micron space via `px = 0.2125 µm/px`. The
  JPEG2000-compressed OME-TIFF is read with Bioconductor's `RBioFormats` (pure
  R); **not** vertically flipped (opposite of MERFISH). If `RBioFormats` is
  missing, the demo is built without the image (coordinates still ship).

# Export to Cerebro (`.crb`)

```r
library(CerebroNexus)
exportFromSeurat(
  object = obj,
  assay = "Spatial",
  slot = "data",
  file = "xenium_mouse_brain.crb",
  experiment_name = "Xenium mouse brain (CTX+HP)",
  organism = "mm",
  groups = c("cluster"),
  nUMI = "nUMI",
  nGene = "nGene"
)

## exportFromSeurat has no image= argument -- the real DAPI morphology channel
## is injected as a second pass: read the .crb back, attach the image to its
## spatial slot, re-save.
crb <- readRDS("xenium_mouse_brain.crb")
sd <- crb$getSpatialData(crb$availableSpatial()[1])
sd$histology_image <- img_uri            # base64 PNG data: URI, native orientation
sd$histology_image_bounds <- img_bounds  # dims * um_per_px, in coordinate space
crb$addSpatialData(crb$availableSpatial()[1], sd)
saveRDS(crb, "xenium_mouse_brain.crb")
```

# Build & launch the Shiny app

```r
library(CerebroNexus)

crb <- "xenium_mouse_brain.crb"

createShinyApp(
  cerebro_data = c("Mouse brain (Xenium)" = crb),
  result_dir   = "my_cerebro_app",
  launch_browser = FALSE
)

# Local:
shiny::runApp("my_cerebro_app")

# shinyapps.io:
# rsconnect::deployApp(appDir = "my_cerebro_app")
```

Open the **Spatial** tab and pick "Mouse brain (Xenium)" from the dataset
switcher.

# Full build code

Produced by `build_xenium()`, one of four functions in
[`data-raw/build_spatial_demos.R`](https://github.com/duocang/CerebroNexus/blob/master/data-raw/build_spatial_demos.R)
(shared download/sampling helpers are defined earlier in that file):

```{r code=local({src <- xfun::read_utf8("../data-raw/build_spatial_demos.R"); s <- grep("^## 4\\. 10x Xenium", src); e <- grep("^## run all", src)[1] - 2; src[s:e]})}
```
