---
title: "Demo dataset: Mouse ileum (MERFISH)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: Mouse ileum (MERFISH)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> One of four real spatial-transcriptomics demos, each on a different platform.
> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

| | |
|---|---|
| **Technology** | MERFISH (single-cell imaging) |
| **Organism / tissue** | mouse / ileum |
| **Source** | Petukhov et al. 2021, via Bioconductor [`MerfishData::MouseIleumPetukhov2021`](https://bioconductor.org/packages/MerfishData) |
| **License** | Bioconductor `MerfishData` / Petukhov et al. terms |
| **Used in** | [Spatial](spatial_analysis.html) tab |
| **Output** | shipped as `inst/extdata/v1.4/demo_spatial_merfish.crb` (~2.0 MB) |

Chosen because it ships a **genuine DAPI tissue mosaic** with cells already in
DAPI pixel coordinates, so the real image aligns to the points almost 1:1.

# Download raw data

```r
BiocManager::install("MerfishData")
spe <- MerfishData::MouseIleumPetukhov2021()   # SpatialExperiment, 241 genes x 5,800 cells
```

# Process in Seurat

The source is a `SpatialExperiment`, not a Seurat object — it is assembled into
one, with a proper `FOV` so extraction runs through the same
`GetTissueCoordinates` path a `LoadVizgen`/`LoadXenium` object would use:

```r
set.seed(42)
cd <- as.data.frame(SummarizedExperiment::colData(spe))
xy <- as.data.frame(SpatialExperiment::spatialCoords(spe))[, c("x", "y")]

## drop the dataset's own discard label
keep <- colnames(spe)[!is.na(cd$leiden_final) & cd$leiden_final != "Removed"]
# then stratified down-sample `keep` to <= 5,000 by cd$leiden_final

mat <- SummarizedExperiment::assay(spe, "counts")[, keep]
obj <- CreateSeuratObject(counts = mat, assay = "Spatial")
obj$cell_type <- factor(as.character(cd[keep, "leiden_final"]))

cents <- CreateCentroids(as.matrix(xy[keep, c("x", "y")]))
obj[["fov"]] <- CreateFOV(list(centroids = cents), type = "centroids", assay = "Spatial")

obj <- NormalizeData(obj) |> FindVariableFeatures() |> ScaleData() |>
  RunPCA(npcs = 30) |> RunUMAP(dims = 1:20)
```

- **sampling**: `leiden_final == "Removed"` dropped, then ≤ 5,000 stratified by
  cell type (here 5,000 of 5,800); 241-gene panel kept whole.
- **cell-type field**: `cell_type` (source `leiden_final` labels — Enterocyte
  tiers, Goblet, Paneth, Smooth Muscle, Myenteric Plexus, ICC, Telocyte, …).
- **embedded image**: **real DAPI mosaic** (9392×5721), stored as
  `histology_image`; **vertically flipped on encode** (`flip_y = TRUE`) — the
  opposite of Xenium (verified by a brightness test; see `spatial.md`).

# Export to Cerebro (`.crb`)

```r
library(CerebroNexus)
exportFromSeurat(
  object = obj,
  assay = "Spatial",
  slot = "data",
  file = "merfish_mouse_ileum.crb",
  experiment_name = "MERFISH mouse ileum",
  organism = "mm",
  groups = c("cell_type"),
  nUMI = "nUMI",
  nGene = "nGene"
)

## exportFromSeurat has no image= argument -- the real DAPI mosaic is injected
## as a second pass: read the .crb back, attach the image to its spatial slot,
## re-save.
crb <- readRDS("merfish_mouse_ileum.crb")
sd <- crb$getSpatialData(crb$availableSpatial()[1])
sd$histology_image <- dapi_uri            # base64 PNG data: URI, flip_y = TRUE
sd$histology_image_bounds <- dapi_bounds  # list(xmin, xmax, ymin, ymax) in coordinate space
crb$addSpatialData(crb$availableSpatial()[1], sd)
saveRDS(crb, "merfish_mouse_ileum.crb")
```

# Build & launch the Shiny app

```r
library(CerebroNexus)

crb <- "merfish_mouse_ileum.crb"

createShinyApp(
  cerebro_data = c("Mouse ileum (MERFISH)" = crb),
  result_dir   = "my_cerebro_app",
  launch_browser = FALSE
)

# Local:
shiny::runApp("my_cerebro_app")

# shinyapps.io:
# rsconnect::deployApp(appDir = "my_cerebro_app")
```

Open the **Spatial** tab and pick "Mouse ileum (MERFISH)" from the dataset
switcher.

# Full build code

Produced by `build_merfish()`, one of four functions in
[`data-raw/build_spatial_demos.R`](https://github.com/duocang/CerebroNexus/blob/master/data-raw/build_spatial_demos.R)
(shared download/sampling helpers are defined earlier in that file):

```{r code=local({src <- xfun::read_utf8("../data-raw/build_spatial_demos.R"); s <- grep("^## 3\\. MERFISH", src); e <- grep("^## 4\\. 10x Xenium", src)[1] - 2; src[s:e]})}
```
