---
title: "Demo dataset: PBMC - Full (T+B)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: PBMC - Full (T+B)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

| | |
|---|---|
| **Technology** | 10x Chromium 5′ V(D)J (scRNA-seq + TCR/BCR) |
| **Organism / tissue** | human / PBMC, healthy donor |
| **Source** | 10x Genomics public dataset `vdj_v1_hs_pbmc3` (Cell Ranger 3.1.0) |
| **License** | 10x Genomics public dataset terms |
| **Used in** | [Immune repertoire](immune_repertoire_analysis.html) tab · [Trajectory](trajectory_analysis.html) tab · [Linked views](../) (default dataset) |
| **Output** | shipped as `inst/extdata/v1.4/demo_full_tcr_bcr.crb` (~1.0 MB) |

This is the one dataset that carries **three** modalities at once: gene
expression, immune repertoire (TCR + BCR), and — via
[the trajectory build step](demo_dataset_trajectory.html) — a pseudotime
trajectory over its B cells.

**A different pipeline shape.** Unlike the four spatial demos (raw counts →
fresh Seurat object → `.crb`), this build starts from an **already-processed**
`.crb` (`example.crb`, itself built from a standard Seurat pipeline documented
separately) and adds the immune-repertoire modality by operating directly on
the Cerebro object structure. So "process" and "export" below are really one
step: read the raw VDJ contigs, turn them into clonotypes, and inject them into
a subset of the base object.

# Download raw data

```bash
mkdir -p data-raw/vdj_10x
BASE=https://cf.10xgenomics.com/samples/cell-vdj/3.1.0/vdj_v1_hs_pbmc3
curl -fL -o data-raw/vdj_10x/pbmc3_t_contig.csv \
  "$BASE/vdj_v1_hs_pbmc3_t_filtered_contig_annotations.csv"
curl -fL -o data-raw/vdj_10x/pbmc3_b_contig.csv \
  "$BASE/vdj_v1_hs_pbmc3_b_filtered_contig_annotations.csv"
```

# Process the repertoire (scRepertoire) & assemble the Cerebro object

```r
library(scRepertoire)
library(CerebroNexus)

pool_from <- function(csv, kind) {
  raw <- read.csv(csv, stringsAsFactors = FALSE)
  contigs <- loadContigs(raw, format = "10X")
  combined <- if (kind == "TCR") combineTCR(contigs) else combineBCR(contigs, threshold = 0.85)
  df <- do.call(rbind, combined)
  unique(df[, c("barcode", "CTgene", "CTnt", "CTaa", "CTstrict")])
}
pool_tcr <- pool_from("data-raw/vdj_10x/pbmc3_t_contig.csv", "TCR")
pool_bcr <- pool_from("data-raw/vdj_10x/pbmc3_b_contig.csv", "BCR")

## The base object (example.crb) already contains only the T+B+Mono subset
## this demo needs (1,476 cells) -- no further cell filtering happens here.
old <- readRDS("inst/extdata/v1.4/example.crb")
full_meta <- old$getMetaData()
```

- **sampling**: the base object (`example.crb`) is already the T + B + Mono
  subset (1,476 cells); no additional filtering is applied at this step.
- **clonotypes**: assigned **only** to cells of the matching lineage — TCR
  pool → T cells, BCR pool → B cells — so the repertoire is biologically
  plausible rather than random noise.
- **cell-type field**: `cell_type` (inherited from `example.crb`).

# Export to Cerebro (`.crb`)

The Cerebro object is rebuilt from the base object's fields (a fresh
`Cerebro_v1.3` with `expression`/`meta_data`/`projections`/`groups` copied
over), and the clonotype pools are split per sample and written into the
object's own `immune_repertoire` field — that field, not a nested key under
`technical_info`, is what the app reads:

```r
new <- Cerebro_v1.3$new()
# ... copy expression / meta_data / projections / groups from `old` ...

ir_df <- rbind(pool_tcr, pool_bcr)   # both pools, cell-type-constrained above
new$immune_repertoire <- split_ir_by_sample(ir_df, new$getMetaData())

saveRDS(new, "pbmc_tcr_bcr.crb")
```

# Build & launch the Shiny app

```r
library(CerebroNexus)

crb <- "pbmc_tcr_bcr.crb"

createShinyApp(
  cerebro_data = c("PBMC - Full (T+B)" = crb),
  result_dir   = "my_cerebro_app",
  launch_browser = FALSE
)

# Local:
shiny::runApp("my_cerebro_app")

# shinyapps.io:
# rsconnect::deployApp(appDir = "my_cerebro_app")
```

Open the **Immune repertoire** tab, or [Linked views](../) (this is its
default dataset).

# Full build code

```{r code=xfun::read_utf8('../data-raw/build_ir_demos.R')}
```
