---
title: "Demo dataset: Mouse brain (Trekker)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: Mouse brain (Trekker)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

| | |
|---|---|
| **Technology** | Trekker Single-Cell Spatial Mapping (Curio Bioscience / Takara Bio; snRNA + bead spatial barcodes). Real single nuclei × whole transcriptome, positions inferred from bead barcodes, no histology image. |
| **Organism / tissue** | mouse / brain, coronal section (`TrekkerU_C`, tile `U0016_004`) |
| **Source** | official Curio/Takara Trekker example bundle `Mouse_Brain_TrekkerU_C_Sept2025` (~1.3 GB). **Registration required — not a public download**; obtained through a Curio/Takara account. **Not redistributable here**; only the derived, down-sampled `.crb` ships. |
| **License** | Curio/Takara example data, access-gated. Shipped `.crb` is a small derived subset. |
| **Used in** | app's bespoke **Trekker** tab · [Linked views](../) (physical space) · [Trekker spatial mapping](trekker_spatial_mapping.html) tutorial |
| **Output** | shipped as `inst/extdata/v1.4/demo_trekker.crb` (~4.7 MB) |

# Download raw data

```bash
# 1. Register / request access to the Trekker example data (Curio / Takara account).
# 2. Download Mouse_Brain_TrekkerU_C_Sept2025.tar.gz and extract:
tar -xzf Mouse_Brain_TrekkerU_C_Sept2025.tar.gz
# 3. Point the builder at the bundle's output/ dir:
export TREKKER_OUTPUT_DIR=/path/to/Mouse_Brain_TrekkerU_C_Sept2025/output
```

# Process in Seurat

```r
message("reading ConfPositioned Seurat RDS ...")
seu <- readRDS(file.path(Sys.getenv("TREKKER_OUTPUT_DIR"), "..._ConfPositioned_seurat_spatial.rds"))

## find the 50 nuclei carrying an official positioning-evidence image
ev_bc <- names(evidence_images)
message("positioning-evidence nuclei found in object: ", length(ev_bc))

## stratified down-sample by Louvain cluster, force-including all evidence nuclei
strat <- integer(0)
for (lv in sort(unique(seu$seurat_clusters))) {
  w <- which(seu$seurat_clusters == lv)
  k <- max(1L, round(2532 * length(w) / ncol(seu)))
  strat <- c(strat, sample(w, min(k, length(w))))
}
idx <- sort(unique(c(strat, match(ev_bc, colnames(seu)))))
sub <- seu[, idx]
message("sub-sampled nuclei: ", length(idx), " (of ", ncol(seu), ")")

sub$celltype <- CELLTYPE_BY_CLUSTER[as.integer(as.character(sub$seurat_clusters)) + 1L]
sub$cluster <- factor(as.integer(as.character(sub$seurat_clusters)))
sub <- DietSeurat(sub, assays = "SCT", dimreducs = "umap")   # lean: drop scale.data, pca, SPATIAL
```

- **sampling**: 2,532 nuclei stratified by Louvain cluster out of 7,420
  confidently-positioned, **plus the 50 nuclei carrying official
  positioning-evidence images force-included**. Whole transcriptome kept (all
  21,374 genes — the platform's differentiator); nuclei are down-sampled, not
  genes.
- **coordinates**: taken from the vendor **Location CSV** (the coordinate
  authority) — never `@images` (transposed) or the `SPATIAL` reduction
  (y-mirrored).
- **embedded image**: **none** — instead the `.crb` embeds, in its `trekker`
  slot, the 50 official per-nucleus evidence JPEGs (base64, down-scaled) so
  the whole demo is self-contained.

# Export to Cerebro (`.crb`)

```r
message("exportFromSeurat (whole transcriptome, ", nrow(sub), " genes) ...")
exportFromSeurat(
  sub,
  assay = "SCT",
  slot = "data",
  file = "trekker_mouse_brain.crb",
  experiment_name = "Trekker mouse brain (TrekkerU_C, demo)",
  organism = "mm",
  groups = c("cluster", "celltype"),
  main_group = "celltype",
  nUMI = "nUMI",
  nGene = "nGene"
)

## exportFromSeurat only knows Seurat objects, so the Trekker-specific payload
## (physical coordinates, vendor QC, positioning evidence) is injected as a
## second pass: read the .crb back, add the slot, re-save.
crb <- readRDS("trekker_mouse_brain.crb")
crb$addTrekker(list(
  x = ..., y = ...,                    # Location CSV coordinates
  fields = list(...),                  # spatial purity, cross-space concordance, position confidence
  conf = list(...),                    # per-nucleus bead statistics behind the confidence
  moran = moran,                       # upstream Moran's I table
  evidence = evidence,                 # 50 base64 positioning-evidence JPEGs
  qc_examples = qc_examples
))
saveRDS(crb, "trekker_mouse_brain.crb", compress = "xz")
```

- **groups**: `cluster` (Louvain) and `celltype` (marker-based annotation per
  cluster) — `celltype` is the `main_group`, so it is what the app colours by
  first.

# Build & launch the Shiny app

```r
library(CerebroNexus)

crb <- "trekker_mouse_brain.crb"

createShinyApp(
  cerebro_data = c("Mouse brain (Trekker)" = crb),
  result_dir   = "my_cerebro_app",
  launch_browser = FALSE
)

# Local:
shiny::runApp("my_cerebro_app")

# shinyapps.io:
# rsconnect::deployApp(appDir = "my_cerebro_app")
```

Open the **Trekker** tab, or [Linked views](../) — the Trekker physical map is
a first-class space there too.

# Full build code

```{r code=xfun::read_utf8('../data-raw/build_trekker_demo.R')}
```
