---
title: "Demo dataset: Mouse brain (10x Visium)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: Mouse brain (10x Visium)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> One of four real spatial-transcriptomics demos, each on a different platform.
> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

| | |
|---|---|
| **Technology** | 10x Visium (spot, 55 µm) |
| **Organism / tissue** | mouse / brain, sagittal anterior section |
| **Source** | 10x Genomics Visium Mouse Brain (Sagittal Anterior), via `stxBrain.SeuratData` |
| **License** | 10x Genomics public dataset terms (freely redistributable) |
| **Used in** | [Spatial](spatial_analysis.html) tab · [Linked views](../) |
| **Output** | shipped as `inst/extdata/v1.4/demo_spatial_visium.crb` (~6.1 MB) + `demo_spatial_visium_he.png` (~0.5 MB) |

This is the deliberate live example of the **external-image** path: unlike the
other three spatial demos, the histology image is written to a standalone PNG
and loaded via `spatial_images` in `inst/app.R`, not embedded in the `.crb`.

# Download raw data

```r
install.packages("SeuratData")   # or remotes::install_github("satijalab/seurat-data")
SeuratData::InstallData("stxBrain")
obj <- SeuratData::LoadData("stxBrain", type = "anterior1")
```

# Process in Seurat

The loaded object is already a Seurat `VisiumV2` object (Seurat 5). The source
ships no cell-type labels, so a Louvain clustering becomes the `cell-type`
field:

```r
set.seed(42)
assay <- "Spatial"
obj <- NormalizeData(obj, assay = assay) |>
  FindVariableFeatures(assay = assay, nfeatures = 2000) |>
  ScaleData(assay = assay) |>
  RunPCA(assay = assay, npcs = 30) |>
  RunUMAP(dims = 1:20) |>
  FindNeighbors(dims = 1:20) |>
  FindClusters(resolution = 0.5)
obj$cluster <- factor(paste0("C", as.integer(obj$seurat_clusters)))
```

- **sampling**: ≤ 5,000 spots stratified by cluster (here 2,696 spots, the full
  anterior1 section); expression trimmed to 1,200 genes (variable + top-expressed)
  to keep the `.crb` small.
- **cell-type field**: `cluster` (`FindClusters(resolution = 0.5)`).
- **embedded image**: **none — an EXTERNAL image instead.** The real low-res H&E
  (599×600×3) is written to `demo_spatial_visium_he.png` and loaded via
  `spatial_images` in `inst/app.R`, displayed with `spatial_images_flip_y = TRUE`.

# Export to Cerebro (`.crb`)

```r
library(CerebroNexus)
exportFromSeurat(
  object = obj,
  assay = "Spatial",
  slot = "data",
  file = "visium_mouse_brain.crb",
  experiment_name = "Visium mouse brain (anterior)",
  organism = "mm",
  groups = c("cluster"),
  nUMI = "nUMI",
  nGene = "nGene"
)
```

# Build & launch the Shiny app

```r
library(CerebroNexus)

crb <- "visium_mouse_brain.crb"

createShinyApp(
  cerebro_data = c("Mouse brain (Visium)" = crb),
  result_dir   = "my_cerebro_app",
  launch_browser = FALSE
)

# Local:
shiny::runApp("my_cerebro_app")

# shinyapps.io:
# rsconnect::deployApp(appDir = "my_cerebro_app")
```

Open the **Spatial** tab and pick "Mouse brain (Visium)" from the dataset
switcher.

# Full build code

Produced by `build_visium()`, one of four functions in
[`data-raw/build_spatial_demos.R`](https://github.com/duocang/CerebroNexus/blob/master/data-raw/build_spatial_demos.R)
(shared download/sampling helpers are defined earlier in that file):

```{r code=local({src <- xfun::read_utf8("../data-raw/build_spatial_demos.R"); s <- grep("^## 1\\. 10x Visium", src); e <- grep("^## 2\\. Slide-seq", src)[1] - 2; src[s:e]})}
```
