---
title: "Demo dataset: Trajectory (B-cell maturation)"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Demo dataset: Trajectory (B-cell maturation)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> See [Demo datasets: overview](demo_datasets.html) for the full lineup.

# Overview

| | |
|---|---|
| **Method** | monocle2, `DDRTree` pseudotime |
| **Cells** | 915 B cells from [`demo_full_tcr_bcr.crb`](demo_dataset_pbmc_tcr_bcr.html) |
| **Source** | derived entirely from the PBMC demo — **no new download** |
| **Used in** | [Trajectory](trajectory_analysis.html) tab |
| **Output** | overwrites the `trajectory` slot inside `demo_full_tcr_bcr.crb` |

**Not a separate `.crb`.** This is the one demo where "download raw data" is a
no-op: the trajectory is computed on cells that already live inside the PBMC
demo and written back into it, so a single dataset ends up demonstrating TCR +
BCR + trajectory together.

**Honest scope.** These are peripheral-blood B cells, not a bone-marrow
developmental lineage — the trajectory is *illustrative* of the pseudotime
feature, not a biological claim about B-cell ontogeny.

# "Download" raw data

None — this step re-reads the already-built `.crb` from the PBMC dataset page:

```r
crb <- readRDS("pbmc_tcr_bcr.crb")
```

# Process: subset B cells and run monocle2

```r
library(monocle)

md <- crb$getMetaData()
b_idx <- which(md$cell_type == "B cells")
expr_b <- crb$getExpressionMatrix()[, b_idx, drop = FALSE]

## Already log-normalized -> Gaussian family, not monocle2's default negbinomial
cds <- newCellDataSet(
  as(as.matrix(expr_b), "sparseMatrix"),
  phenoData = new("AnnotatedDataFrame", data = data.frame(cell_barcode = md$cell_barcode[b_idx])),
  featureData = new("AnnotatedDataFrame", data = data.frame(gene_short_name = rownames(expr_b))),
  expressionFamily = VGAM::uninormal()
)

ordering_genes <- names(sort(apply(as.matrix(expr_b), 1, var), decreasing = TRUE))[1:500]
cds <- setOrderingFilter(cds, ordering_genes)
cds <- reduceDimension(cds, max_components = 2, reduction_method = "DDRTree",
                        norm_method = "none", pseudo_expr = 0)
cds <- orderCells(cds)
```

- **sampling**: `set.seed(42)`; all 915 B cells kept, no down-sampling.
- **ordering**: top 500 genes by variance.
- Two small compatibility shims patch two `monocle2` internal functions that
  call igraph APIs removed in current igraph (`graph.dfs`/`nei` → `dfs`/`.nei`)
  — exact drop-in replacements, not a reimplementation of the ordering logic.
  See the full script below for the patch code and its self-checks.

# Export: write the trajectory back into the `.crb`

No new file — the pseudotime coordinates and the minimum-spanning-tree edges
are added to the *existing* object's `trajectory` slot and the same `.crb` is
overwritten:

```r
crb$addTrajectory(
  "monocle2",
  "B_cell_maturation",
  list(
    meta  = data.frame(DR_1 = ..., DR_2 = ..., pseudotime = ..., state = ...),
    edges = data.frame(source = ..., target = ..., source_dim_1 = ..., ...)
  )
)
saveRDS(crb, "pbmc_tcr_bcr.crb")
```

# Build & launch the Shiny app

Same `.crb` as the [PBMC - Full (T+B)](demo_dataset_pbmc_tcr_bcr.html) demo —
the trajectory just travels with it:

```r
library(CerebroNexus)

crb <- "pbmc_tcr_bcr.crb"

createShinyApp(
  cerebro_data = c("PBMC - Full (T+B)" = crb),
  result_dir   = "my_cerebro_app",
  launch_browser = FALSE
)

# Local:
shiny::runApp("my_cerebro_app")

# shinyapps.io:
# rsconnect::deployApp(appDir = "my_cerebro_app")
```

Open the **Trajectory** tab.

# Full build code

```{r code=xfun::read_utf8('../data-raw/build_trajectory_demo.R')}
```
