---
title: "Extra Material"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Extra Material}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

The **Extra material** tab provides a space for custom tables or plots
bundled alongside your single-cell data. Common use cases include:

- Cell type annotation results (e.g., SingleR scores)
- Custom QC summary tables
- Publication-ready figures
- Any additional analysis you want to share with collaborators

The tab appears _conditionally_ when the `.crb` file contains extra material
or when an app is built with external tables. You control what appears, so the
interface stays clean when no extra content is needed.

# Quick start

```{r, eval = FALSE}
library(CerebroNexus)
launchCerebro()
```

1. Launch CerebroNexus and load a `.crb` file with extra material, or open a
   generated app with external tables
2. If extra material is present, **Extra material** appears in the sidebar
3. Select a category (`tables` or `plots`)
4. For tables, choose a File and then a Sheet when more than one is available
5. Choose a specific plot to view

# Content categories

**Tables**: any `data.frame`. Rendered as interactive DT tables with search,
sort, and download.

**Plots**: any `ggplot2` object. Displayed at original resolution with
download support.

# Embedding content

Use the `extra_material` parameter in `exportFromSeurat()`:

```{r, eval = FALSE}
exportFromSeurat(seurat_object,
  file = "my_data.crb",
  extra_material = list(
    tables = list(SingleR_results = annotation_df),
    plots  = list(umap_overview = umap_plot)
  )
)
```

# Add external table files when building an app

`createShinyApp()` can freeze CSV, TSV, TXT, XLS, XLSX, and XLSM files into a
generated app. The app reads the files during the build, so deployed Viewers
do not need access to the original paths. The names in `extra_tables` become
the **File** choices in the Viewer.

A delimited file contributes one table. An Excel workbook contributes every
non-empty sheet. Use `extra_tables_sheets` only to rename Excel sheets; it
never filters sheets out. If a workbook has one file or a file has one sheet,
the corresponding selector is hidden.

## A CSV table

```{r, eval = FALSE}
createShinyApp(
  cerebro_data = c("Study" = "study.crb"),
  extra_tables = c("QC summary" = "supplements/qc.csv"),
  result_dir = "my-app"
)
```

This displays `QC summary` as one File containing one table. No
`extra_tables_sheets` entry is needed (or allowed) for a CSV file.

## Two Excel workbooks, with one sheet renamed

```{r, eval = FALSE}
createShinyApp(
  cerebro_data = c("Study" = "study.crb"),
  extra_tables = list(
    "Clinical workbook" = "supplements/clinical.xlsx",
    "Marker workbook" = "supplements/markers.xlsx",
    "QC summary" = "supplements/qc.csv"
  ),
  extra_tables_sheets = list(
    "Clinical workbook" = list(
      "Patient characteristics" = "Patients"
    )
  ),
  result_dir = "my-app"
)
```

For example, if `clinical.xlsx` contains `Patients`, `Visits`, and `Samples`,
the Viewer shows the File `Clinical workbook` and the Sheets `Patient
characteristics`, `Visits`, and `Samples`. Only `Patients` was renamed; the
other non-empty sheets remain available. The original file paths are not
stored in the generated app configuration.

# See also

`vignette("export_and_visualize_custom_tables_and_plots")` for a detailed
walkthrough with examples.
