---
title: "Host Cerebro on shinyapps.io"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Host Cerebro on shinyapps.io}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

In this vignette, we explain how one can host Cerebro on [shinyapps.io](https://www.shinyapps.io).
For a more detailed tutorial on how to host Shiny apps on shinyapps.io, take a look here: <https://shiny.rstudio.com/articles/shinyapps.html>

[shinyapps.io](https://www.shinyapps.io) is a great place to host Shiny apps and make them accessible for other users.
You will need an account on the website.
The free account comes with a limit on (1) the number of applications that can be hosted in parallel and (2) the number of active hours, but it is enough to get some experience and figure out if this service fits your purpose.

> **Bundle first.** The instructions below build the deployable directory with
> `createShinyApp()`, the same function used in
> [Create a self-contained Shiny app](create_a_self_contained_shiny_app.html).
> Read that vignette first if you haven't already — this one just adds the
> `rsconnect` upload step on top.

# Setup

To upload Cerebro to shinyapps.io, we will use the [`rsconnect`](https://cran.r-project.org/web/packages/rsconnect/index.html) package, therefore it needs to be installed first.
Of course, we also need the `CerebroNexus` package.

```{r eval=FALSE}
if ( 'rsconnect' %in% installed.packages() == FALSE ) install.packages('rsconnect')
if ( 'CerebroNexus' %in% installed.packages() == FALSE ) install.packages('mihem/CerebroNexus')

library(rsconnect)
```

Then, you need to provide `rsconnect` with your shinyapps.io account info.
On the shinyapps.io website, in the left navigation bar, go to "Account" and then "Tokens".
A token should have already been generated for you.
Take the token and secret, your account name, and run the command below.

```{r eval=FALSE}
rsconnect::setAccountInfo(name="<ACCOUNT>", token="<TOKEN>", secret="<SECRET>")
```

# Build the app bundle

`createShinyApp()` writes a complete, deployable directory — the Shiny
sources, your `.crb` file(s), the `inst/extdata/` reference files, and an
`app.R` that wires everything together. That directory is exactly what
`rsconnect::deployApp()` expects, so there is no manual file-copying or
hand-written `app.R` needed.

```{r eval=FALSE}
library(CerebroNexus)

crb <- system.file("extdata/v1.4/example.crb", package = "CerebroNexus")

createShinyApp(
  cerebro_data = c("PBMC example" = crb),
  result_dir   = "~/test_cerebro_shinyapps",
  launch_browser = FALSE
)
```

See [Create a self-contained Shiny app](create_a_self_contained_shiny_app.html)
for the full set of `createShinyApp()` arguments (multiple data sets, a
custom welcome message, `.h5`/`.bpcells` expression backends, and so on).

# Deploy app

Now, it is time to upload the app to shinyapps.io.
With `rsconnect`, that can be quickly done using the code below.

Since we have a dependency from [Bioconductor](http://bioconductor.org), we have to add the current release to the repositories using the `setRepositories()` command.
When doing so, you will be asked which repositories should be used in this session.
For what we are doing here, selecting `CRAN` and `BioC software` should be sufficient.

```{r eval=FALSE}
setRepositories(addURLs = c(BioC = "https://bioconductor.org/packages/release/bioc"))

rsconnect::deployApp('~/test_cerebro_shinyapps/', appName = 'Cerebro')
```

Uploading and preparing the app might take a few minutes, but once it finishes successfully, Cerebro should open in your browser.
Using the shinyapps.io dashboard, you can manage the app, check usage metrics and logs, etc.
An important parameter is the "Instance size", which controls the available memory and can be controlled in the "Settings" tab of the dashboard.
The larger the data set you want to upload to Cerebro, the more memory will be necessary.

# See also

* [Create a self-contained Shiny app](create_a_self_contained_shiny_app.html)
* [Launch Cerebro with pre-loaded data set](launch_cerebro_with_pre-loaded_data_set.html)
* [Control access to Cerebro with a login page](control_access_to_cerebro_with_a_login_page.html)

# Known issues

* Exporting plots to PDF through the `export to PDF` buttons effectively does not work at the moment. That is because it will trigger a file selection dialog on the server side, not allowing the user to specify a location on their local machine.
