---
title: "Reproduce the 1M-cell Viewer benchmark"
author: "CerebroNexus contributors"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Reproduce the 1M-cell Viewer benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Scope

This vignette compares Viewer performance between two Git revisions. It uses
the official 10x Genomics E18 mouse brain matrix, retains exactly 1,000,000
cells, and creates a BPCells-backed Seurat object and Cerebro file.

The benchmark has two layers:

1. `tests/bench/viewer_1m_hot_paths.R` measures filtering, sampling, hover
   preparation and expression access directly in R.
2. `tests/bench/viewer_1m_browser.R` launches the real Shiny Viewer in Chrome,
   waits for the 1M dataset and Overview Canvas, checks that the Canvas contains
   plotted pixels, rejects browser errors, and records Shiny process memory.

The scripts are the executable source of truth. This vignette explains their
arguments and supplies complete shell commands rather than copying their
implementation into a second place.

# Compared revisions

Choose a baseline and candidate revision. The baseline must be an ancestor of
the candidate so that the comparison isolates their intervening changes.

```{bash}
export BASELINE_REV="<baseline revision>"
export CANDIDATE_REV="<candidate revision>"

git rev-parse --verify "$BASELINE_REV^{commit}"
git rev-parse --verify "$CANDIDATE_REV^{commit}"
git merge-base --is-ancestor "$BASELINE_REV" "$CANDIDATE_REV"
```

All commands exit with status zero when the revisions are valid.

# Requirements

The reported run used:

- Apple M1 Pro with 32 GiB RAM;
- macOS 27.0;
- R 4.6.1 on `aarch64`;
- BPCells 0.3.1, Seurat 5.5.1, Shiny 1.14.0 and shinytest2 0.5.1;
- a Chrome or Chromium browser discoverable by shinytest2;
- at least 15 GB of free disk space for the source and reusable outputs.

Run this preflight from the repository root. It only reports missing packages;
it does not install or change anything.

```{bash}
Rscript - <<'RS'
required <- c(
  "BPCells", "devtools", "dplyr", "glue", "magrittr", "Matrix",
  "ps", "Seurat", "SeuratObject", "shiny", "shinytest2"
)
missing <- required[
  !vapply(required, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))
]
if (length(missing)) {
  stop("Missing packages: ", paste(missing, collapse = ", "))
}
cat("Benchmark packages: OK\n")
RS
```

# Prepare isolated comparison trees

Do not switch the working checkout back and forth during a benchmark. Create
two detached Git worktrees so the before and after versions coexist while both
read the same CRB.

Copy and run this block from the repository root:

```{bash}
REPO_ROOT="$(git rev-parse --show-toplevel)"
BENCH_ROOT="$(mktemp -d "${TMPDIR:-/tmp}/cerebronexus-viewer-bench.XXXXXX")"
BEFORE_ROOT="$BENCH_ROOT/before"
AFTER_ROOT="$BENCH_ROOT/after"

git worktree add --detach "$BEFORE_ROOT" "$BASELINE_REV"
git worktree add --detach "$AFTER_ROOT" "$CANDIDATE_REV"

printf 'repository: %s\nbenchmark trees: %s\n' "$REPO_ROOT" "$BENCH_ROOT"
```

The browser benchmark passes a named dataset path to `launchCerebro()`, so
both worktrees remain clean and require no compatibility patch.

# Download and prepare the 1M example

Choose a persistent cache location. Reusing it is important because preparation
takes tens of minutes and the source H5 is about 4.2 GB.

```{bash}
export CEREBRO_LARGE_CACHE="${XDG_CACHE_HOME:-$HOME/.cache}/CerebroNexus-benchmark"
mkdir -p "$CEREBRO_LARGE_CACHE"
df -h "$CEREBRO_LARGE_CACHE"
```

Run the complete download, Seurat sketch workflow and CRB conversion from the
source checkout:

```{bash}
cd "$REPO_ROOT"

/usr/bin/time -l Rscript - <<'RS'
devtools::load_all(".", quiet = TRUE)
result <- .prepareLargeExample(
  size = "1m",
  cache_dir = Sys.getenv("CEREBRO_LARGE_CACHE")
)
print(result)
RS
```

The preparation is resumable at artifact boundaries. A later call reuses a
complete source, Seurat object or CRB instead of downloading and converting it
again. Partial downloads are not published as complete cache files.

Define and validate the resulting paths:

```{bash}
CRB="$CEREBRO_LARGE_CACHE/1m/cerebro/cerebro_mouse_brain_1m.crb"
SIDECAR="${CRB%.crb}.bpcells"

test -f "$CRB"
test -d "$SIDECAR"
du -sh "$CEREBRO_LARGE_CACHE/1m/source" \
  "$CEREBRO_LARGE_CACHE/1m/seurat" \
  "$CEREBRO_LARGE_CACHE/1m/cerebro"
```

The benchmarked artifact contained 1,000,000 cells, 27,998 genes, 33 clusters
and one full 1,000,000-row UMAP. Preparation took 1,342.12 seconds and reached
8.47 GiB maximum resident memory.

| Artifact | Exact bytes |
|---|---:|
| Official source H5 | 4,216,018,749 |
| Reusable Seurat output and BPCells sidecar | 3,778,378,648 |
| CRB and BPCells sidecar | 3,765,141,671 |
| CRB metadata file alone | 28,132,278 |

All of these files remain in the selected cache. The large data are never
added to Git.

# Run the R hot-path benchmark

The command-line interface is:

```text
viewer_1m_hot_paths.R BEFORE_ROOT AFTER_ROOT CRB [REPEATS]
```

Run three repetitions and save the tab-separated result:

```{bash}
Rscript "$REPO_ROOT/tests/bench/viewer_1m_hot_paths.R" \
  "$BEFORE_ROOT" \
  "$AFTER_ROOT" \
  "$CRB" \
  3 \
  > "$BENCH_ROOT/viewer_1m_hot_paths.tsv"

column -t -s $'\t' "$BENCH_ROOT/viewer_1m_hot_paths.tsv"
```

For each operation the script:

1. reads the same 1M-cell CRB and BPCells sidecar;
2. sources the Viewer helpers from the before and after worktrees;
3. warms each operation once;
4. runs garbage collection before each timed repetition;
5. reports the median elapsed time;
6. records one R allocation profile with `Rprofmem()`; and
7. stops immediately if the before and after results violate the operation's
   equality or invariant check.

The measured three-repetition medians were:

| Operation | Scale | Before | After | Time | Before allocation | After allocation | Allocation |
|---|---|---:|---:|---:|---:|---:|---:|
| Full projection selection | 1M cells, all groups, 100% | 222 ms | 4 ms | -98.2% | 206.7 MiB | 11.4 MiB | -94.5% |
| Filtered projection selection | 1M cells, 17/33 clusters, 25% | 356 ms | 30 ms | -91.6% | 170.6 MiB | 43.3 MiB | -74.6% |
| Hover preparation | 1M loaded, 100K displayed, 3 groups | 32,666 ms | 2,385 ms | -92.7% | 1,328.7 MiB | 56.3 MiB | -95.8% |
| Single-gene expression | 1 gene x 1M cells, BPCells | 4,041 ms | 4,073 ms | +0.8% | 154.1 MiB | 169.3 MiB | +9.9% |
| RGB expression | 3 genes x 1M cells | 12,285 ms | 4,067 ms | -66.9% | 461.4 MiB | 238.3 MiB | -48.4% |
| Multi-panel expression | 9 genes x 1M cells | 4,336 ms | 4,272 ms | -1.5% | 562.3 MiB | 463.2 MiB | -17.6% |
| Mean expression | 100 genes x 1M cells | 5,540 ms | 3,978 ms | -28.2% | 1,916.8 MiB | 112.8 MiB | -94.1% |

`Rprofmem()` measures allocations known to R, not all native memory owned by
BPCells or the operating system. The browser benchmark below therefore also
records process RSS. The single-gene result is time-neutral in this run; the
main expression gains come from batching RGB reads and keeping mean expression
backend-native.

# Run the browser loading benchmark

The browser command-line interface is:

```text
viewer_1m_browser.R BEFORE_ROOT AFTER_ROOT CRB [REPEATS] [PERCENT]
```

`PERCENT` is the percentage of cells drawn in Overview and must be between 10
and 100. The reported before/after comparison uses 10, meaning that the real
1M dataset is loaded while 100,000 points are drawn. Run:

```{bash}
Rscript "$REPO_ROOT/tests/bench/viewer_1m_browser.R" \
  "$BEFORE_ROOT" \
  "$AFTER_ROOT" \
  "$CRB" \
  3 \
  10 \
  > "$BENCH_ROOT/viewer_1m_browser.txt"

cat "$BENCH_ROOT/viewer_1m_browser.txt"
```

Each repetition performs the following checks and measurements:

1. generates a temporary app that loads the selected Git revision;
2. launches it through shinytest2 and Chrome;
3. opens Data Info and waits until it reports exactly 1,000,000 cells;
4. records `data_ready_ms`;
5. opens Overview and waits for a non-zero Canvas;
6. samples Canvas pixels and requires at least 1,000 non-white samples;
7. rejects unexpected browser `error`, `fatal` or `unhandled` log messages;
8. records the Shiny R process RSS; and
9. alternates before/after execution order between rounds.

The output has a `RAW` table with every run and a `SUMMARY` table containing
the medians. The three-round result was:

| Metric | Before median | After median | Change |
|---|---:|---:|---:|
| Data ready | 16,315 ms | 15,981 ms | -2.0% |
| Overview Canvas ready | 46,609 ms | 19,102 ms | -59.0% |
| Launch through painted Overview | 62,460 ms | 35,418 ms | -43.3% |
| Shiny process RSS | 3,759.6 MiB | 2,160.3 MiB | -42.5% |

A separate rerun from clean detached checkouts measured 64,033 ms and 3,530.4
MiB before, versus 34,105 ms and 2,699.5 MiB after. Both versions painted the
same 1301-by-594 Canvas with more than 31,000 sampled non-white pixels.

# Validate full 1M-point Canvas rendering

The 10% comparison exercises the intended large-data default. To validate the
chunked Canvas path, run the candidate checkout as both inputs and request 100%
rendering:

```{bash}
Rscript "$REPO_ROOT/tests/bench/viewer_1m_browser.R" \
  "$AFTER_ROOT" \
  "$AFTER_ROOT" \
  "$CRB" \
  2 \
  100 \
  > "$BENCH_ROOT/full_1m_canvas.txt"

cat "$BENCH_ROOT/full_1m_canvas.txt"
```

This is a rendering acceptance check, not a baseline/candidate comparison. The
measured runs painted all 1M UMAP points in 72.4 and 73.5 seconds, using
4,616.7 and 4,368.5 MiB RSS.

# Use the prepared example normally

Launch it directly from the source checkout:

```{r}
devtools::load_all(".")

launchCerebro(
  example_data_size = "1m",
  example_data_dir = Sys.getenv("CEREBRO_LARGE_CACHE"),
  percentage_cells_to_show = 10
)
```

Or package it together with other Cerebro datasets:

```{r}
devtools::load_all(".")

createShinyApp(
  cerebro_data = c("Small dataset" = "/absolute/path/small.crb"),
  example_data_size = "1m",
  example_data_dir = Sys.getenv("CEREBRO_LARGE_CACHE"),
  result_dir = "/absolute/path/cerebro-large-app",
  launch_browser = FALSE
)
```

# Interpret and compare runs

Use the medians, not a single fastest run. Keep the following fixed between
versions:

- CRB and BPCells sidecar;
- display percentage and browser viewport;
- R, package and browser versions;
- machine power mode and competing workloads; and
- number of repetitions.

The first read may populate operating-system disk caches. Alternating browser
order reduces, but does not eliminate, thermal and cache bias. Report the raw
runs alongside medians, and do not describe a sub-percent difference as a
speedup without additional repetitions.

# Clean up comparison worktrees

The cache is intentionally retained. Remove only the temporary Git worktrees
after collecting the result files you need:

```{bash}
git -C "$REPO_ROOT" worktree remove "$BEFORE_ROOT"
git -C "$REPO_ROOT" worktree remove "$AFTER_ROOT"
```

The benchmark result files remain under `$BENCH_ROOT`. Delete that temporary
directory separately only after copying any results you want to keep.
