---
title: "Launch Cerebro with pre-loaded data set"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Launch Cerebro with pre-loaded data set}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`launchCerebro()` can launch with a `.crb` file pre-loaded and prevent the
user from uploading their own data set.
This is useful when you want to share your Cerebro data set on a web server,
e.g. along with a publication, but don't want users to use your web server to
upload their own data sets.

> For a fully portable, self-contained bundle (recommended for most sharing
> scenarios), see [Create a self-contained Shiny app](create_a_self_contained_shiny_app.html)
> instead — `createShinyApp()` produces a directory that already has a
> pre-loaded data set and needs no further `launchCerebro()` call.

# How to do it

It's very simple, all you have to do is provide some parameters to
`launchCerebro()`.
Let's say your data set is stored here: `~/Cerebro/data_sets/pbmc_v3.crb`.
Then, launch Cerebro as shown below:

```{r eval=FALSE}
launchCerebro(
  version = "v1.4",
  crb_file_to_load = "~/Cerebro/data_sets/pbmc_v3.crb",
  mode = "closed"
)
```

That's it.

# Modify welcome page

If you would like to modify the message on the "Load data"/welcome page, e.g. to link to a publication or give an introduction to the data set, you can use the `welcome_message` parameter.
The provided string for that parameter can/should be written in HTML as shown below:

```{r eval=FALSE}
custom_welcome_message <- '<h3 style="text-align: center; margin-top: 0px"><strong>This is a custom welcome message to Cerebro</strong></h3>
  <p style="margin-left: 10px;">This data set belongs to publication XY by Max Mustermann <em>et al.</em>.<br>
  It contains the transcriptomic profiles of PBMC samples taken from mice undergoing treatment F.<br>
  Please contact us at ... if you have any questions.</p>'

launchCerebro(
  version = "v1.4",
  crb_file_to_load = "~/Cerebro/data_sets/pbmc_v3.crb",
  mode = "closed",
  welcome_message = custom_welcome_message
)
```

# See also

* [Host Cerebro on shinyapps.io](host_cerebro_on_shinyapps.html)
* [Control access to Cerebro with a login page](control_access_to_cerebro_with_a_login_page.html)
