---
title: "Million-cell Viewer backend hot paths"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Million-cell Viewer backend hot paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

historical <- data.frame(
  operation = c(
    "Full projection selection",
    "Filtered projection selection",
    "Single-gene expression",
    "RGB expression",
    "Multi-panel expression",
    "Mean expression"
  ),
  scale = c(
    "1M cells, all groups, 100%",
    "1M cells, 17/33 clusters, 25%",
    "1 gene x 1M cells",
    "3 genes x 1M cells",
    "9 genes x 1M cells",
    "100 genes x 1M cells"
  ),
  before_ms = c(226, 345, 4167, 12310, 5327, 5861),
  after_ms = c(31, 28, 4182, 4106, 4702, 4093),
  before_alloc_mib = c(206.7, 170.6, 154.1, 461.4, 562.3, 1916.8),
  after_alloc_mib = c(19.1, 39.5, 169.3, 238.3, 463.2, 112.8),
  check.names = FALSE
)
historical$time_change <- sprintf(
  "%+.1f%%",
  (historical$after_ms / historical$before_ms - 1) * 100
)
historical$allocation_change <- sprintf(
  "%+.1f%%",
  (historical$after_alloc_mib / historical$before_alloc_mib - 1) * 100
)
```

# Purpose

This benchmark isolates the R-side paths that prepare million-cell Viewer data before JSON serialization or browser rendering. Its purpose is to answer a practical question: when a user filters a projection or switches expression colouring, how much server time and temporary R allocation can be removed without changing the CRB schema, the selected cells, or the expression values?

The comparison deliberately stops before Canvas or WebGPU rendering. It measures metadata filtering and sampling, canonical cell-index handling, BPCells expression reads, RGB and multi-panel batching, and backend-native mean expression. Browser payload and renderer work belong to the following Viewer/WebGPU layer.

# Version and data contract

| Role | Revision | Version | Contract |
|---|---|---|---|
| Before | `27303f21` | CerebroNexus 4.5.0 | Thin CRB v1, qs2 default, canonical BPCells cell index |
| After | current branch | CerebroNexus 4.5.1 | Same CRB contract plus backend hot-path and canonical-index optimizations |

Both sides read the same artifact under `large-examples/1m/cerebro/`. `readCerebro()` hydrates the qs2 or RDS Thin CRB, opens the unchanged BPCells sidecar, verifies the `cell_names` checksum, and restores omitted barcode and projection row names. The benchmark does not create a second data layout.

```text
large-examples/1m/
├── source/
├── seurat/
└── cerebro/
    ├── cerebro_mouse_brain_1m.crb
    └── cerebro_mouse_brain_1m.bpcells/
```

# What changed

Projection filtering now returns original integer row indices directly instead of copying the complete metadata into a temporary data frame. Expression access keeps those canonical indices through `getExpressionRow()` and `getExpressionBlock()`, so a million indices are no longer converted to barcode strings and matched back to the same BPCells columns. RGB reads are batched, multi-panel extraction avoids a full transpose, and mean expression stays in the backend-native matrix representation.

The output is unchanged: sampling still uses the same seed and eligible cells, requested order is preserved, missing names still fail explicitly, and character barcode callers remain supported.

# Recorded historical result

The following measurements were collected on 2026-09-11 for the original `69893a2b` baseline and backend candidate `8f803742`. They are retained to show the observed direction and user-visible scale. They are not presented as measurements of the rebased 4.5.0-to-4.5.1 comparison; the canonical-index follow-up requires a new run.

```{r historical-table}
knitr::kable(
  historical[, c(
    "operation",
    "scale",
    "before_ms",
    "after_ms",
    "time_change",
    "before_alloc_mib",
    "after_alloc_mib",
    "allocation_change"
  )],
  col.names = c(
    "Operation",
    "Scale",
    "Before (ms)",
    "After (ms)",
    "Time",
    "Before allocation (MiB)",
    "After allocation (MiB)",
    "Allocation"
  ),
  row.names = FALSE,
  digits = 1
)
```

In direct terms, full projection filtering fell from 226 to 31 ms and filtered selection from 345 to 28 ms, moving both operations from a visible pause toward immediate feedback. RGB expression fell from 12.31 to 4.11 seconds, saving about 8.2 seconds per change. Mean expression fell from 5.86 to 4.09 seconds while cumulative R allocation dropped from 1,916.8 to 112.8 MiB. The single-gene result remained flat and allocated slightly more memory; that result motivated the 4.5.1 canonical-index follow-up.

One explanatory pass through all six operations fell from 28.236 to 17.142 seconds and from 3,471.9 to 1,042.2 MiB of cumulative R allocations. That is 39.3% less elapsed time and 70.0% less allocation, but it is not an end-to-end Viewer timing because the operations have different UI frequencies and the total excludes transport and rendering.

# Resource interpretation

`before_ms` and `after_ms` are warm-cache wall-clock medians. `before_alloc_mib` and `after_alloc_mib` are cumulative allocations recorded by `Rprofmem`; they are not peak resident memory. The preparation host reached 8.47 GiB maximum resident memory while creating the reusable 1M artifacts, but that preparation cost is outside the interactive hot-path comparison.

| Resource | Recorded value |
|---|---:|
| Cells | 1,000,000 |
| Genes | 27,998 |
| Clusters | 33 |
| End-to-end data preparation | 1,342.12 s |
| Preparation maximum resident memory | 8.47 GiB |
| Official source H5 | 4,216,018,749 bytes |
| Reusable Seurat output | 3,778,378,648 bytes |
| CRB plus BPCells output | 3,765,141,671 bytes |
| Thin qs2 CRB control payload | 13.21 MiB |

# Reproduce the comparison

The wrapper defaults to the exact Thin CRB baseline and current checkout. It creates detached temporary worktrees, records both full SHAs, reuses complete artifacts, and writes a tab-separated before/after table. Existing 1M data are not downloaded or regenerated when the expected files and sidecar are complete.

```{bash reproduce, eval=FALSE}
export CEREBRO_LARGE_CACHE=/path/to/cerebro-large-examples
tests/bench/run_viewer_1m_benchmark.sh 27303f21 HEAD tests/bench/scratch/backend-4.5.1
```

The output directory contains the exact comparison identity and measured rows:

```{bash inspect-output, eval=FALSE}
column -t -s $'\t' tests/bench/scratch/backend-4.5.1/run_manifest.tsv
column -t -s $'\t' tests/bench/scratch/backend-4.5.1/hot_paths.tsv
```

To reuse a prepared CRB directly, run the lower-level benchmark with explicit source roots. The CRB may use qs2 or RDS because the script loads it through `readCerebro()`.

```{bash lower-level, eval=FALSE}
Rscript tests/bench/viewer_1m_hot_paths.R /path/to/before /path/to/after /path/to/cerebro_mouse_brain_1m.crb 3
```

The current benchmark records a separate barcode-to-index row before the expression workloads. This exposes the work removed by retaining canonical indices instead of hiding it inside a BPCells read.

# Evidence policy

Published values must come from `hot_paths.tsv` together with its `run_manifest.tsv`; revision labels copied by hand are insufficient. A row is accepted only when its before/after correctness check passes. Cold-cache results, browser transfer, JSON parsing, Canvas/WebGPU rendering, GPU memory, and end-to-end interaction latency are intentionally outside this backend-only benchmark and must not be inferred from it.
