---
title: "Responsive Shiny workloads with mirai"
author: "CerebroNexus contributors"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
vignette: >
  %\VignetteIndexEntry{Responsive Shiny workloads with mirai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Why this runtime exists

The `master` viewer performs scientific calculations, file reads, plot export,
and reactive orchestration in the same R process. Shiny serializes work within
that process: while one handler constructs a motif graph or Moran statistic,
the process cannot service another input or session.

The optimized runtime keeps Shiny as the owner of sessions, reactive state,
validation, file-backed matrices, and final rendering. It moves pure,
serializable work to bounded `mirai` daemons over NNG. This targets response
latency, not merely task wall time: a task may take the same time in a daemon,
but the Shiny event loop remains available.

# Before and after

| Area | `master` | Optimized branch |
|---|---|---|
| Runtime | All work in the Shiny R process | One bounded `mirai` compute profile, dispatcher, timeout, and queue-memory limit |
| Rapid input changes | Earlier computations continue and may finish late | Latest-wins cancellation plus generation tokens discard stale results |
| HLA motif graph | Distance graph and layout block the session | Plain segment tables are processed in a daemon; finalized graphs retain the existing Shiny cache |
| scRepertoire | Package loading and seven expensive plot/table calls run in the main process | Namespace loading and abundance, diversity, homeostasis, compare, overlap, rarefaction, and size-distribution calls run in daemons |
| Spatial | Moran's I builds a distance matrix in the main process | Only `x`, `y`, and expression vectors cross the boundary; Moran's I runs in a daemon |
| Trajectory | Projection preparation and density estimation block reactive delivery | Immutable data frames/vectors are prepared in daemons |
| Gene expression | RGB mode can read one gene at a time; means are calculated in the main process | One requested genes-by-cells slice is read, then RGB, panels, and means are prepared in a daemon |
| Linked Views | Every session rebuilds the same full bundle; genes are fetched separately | Immutable bundles are shared across sessions; one multi-gene slice feeds asynchronous colour preparation |
| Extra material | First access calls `readRDS()` and sanitizes the table synchronously | External tables are read and formula-sanitized in a daemon, then shared across sessions |
| First browser paint | Hidden controls and nine capability checks join the first reactive flush | The configured CRB is preloaded once, hidden controls and capability checks wait until after the active page paints |
| First page activation | Every projection update waits for its full debounce; opening Projection also builds hidden Linked Views | The first value bypasses debounce, dedicated pages request only their own payload, and Linked Views warms only after first paint while no dedicated canvas is active |
| Linked Views definitions | Bundle and configuration functions are parsed again for every session | Pure definitions are parsed once per R process and rebound to each session |
| CRB loading | Every browser session deserializes the same CRB | The default CRB is preloaded before sessions connect; the live R6/file-backed object remains local and is shared read-only across sessions |
| PDF export | `ggsave()` blocks the event handler | Overview, expression, spatial, and trajectory PDFs are written in a daemon and report completion asynchronously |
| HLA ZIP export | CSV creation and compression start on click in the main process | The archive is prepared in a daemon after the HLA page becomes active; download writes ready bytes |
| App forms | No daemon lifecycle | Installed, source, and generated standalone apps initialize the same runtime, start daemons after first paint/first task, and stop only owned daemons |

# Process boundary

The important boundary is deliberately narrow:

```text
browser input -> Shiny validates/snapshots -> mirai daemon computes
              <- plain R result           <-
              -> Shiny renders/sends result
```

The following objects never cross NNG: `session`, reactive expressions, the
live Cerebro R6 object, HDF5Array/BPCells handles, and graphics devices. A
file-backed expression request is sliced once in the main process because its
handle belongs there; only the small materialized matrix is sent. CRB loading
also remains local for the same reason. Process-level preload/read-only caching
removes that cost from the first reactive flush and later concurrent sessions
without unsafe serialization.

# Runtime configuration

The default uses at most two workers, a 256 MB dispatcher queue, and a five
minute task timeout. `launchCerebro()` accepts overrides:

```{r config, eval=FALSE}
launchCerebro(
  crb_file_to_load = "dataset.crb",
  mirai_options = list(
    workers = 2L,
    queue_memory_mb = 256,
    timeout_ms = 300000L,
    compute = "cerebro"
  )
)
```

Set `enabled = FALSE` to execute the same worker functions synchronously. This
is useful for diagnosis and provides deterministic parity testing; production
apps should normally keep the default enabled runtime.

Installed packages receive `mirai` and `promises` through `Imports`. A source or
standalone checkout with either package missing now emits one warning and uses
the same synchronous workers instead of terminating the Shiny process. Install
the declared dependencies to restore asynchronous execution.

Daemon creation is lazy. Configuration and ownership are established when the
app starts, while owned processes start after the first browser paint or when a
task is submitted first. Thus daemon startup does not delay the initial Shiny
response.

# Reproducible benchmark

The repository includes `benchmarks/mirai_runtime_benchmark.R`. It compares
the unchanged scientific worker synchronously (the `master` execution model)
with a warm one-daemon runtime, verifies scientific equality, and measures a
zero-delay `later` callback as an event-loop responsiveness probe.

Run it from the repository root:

```{r benchmark-command, eval=FALSE}
system2(
  "Rscript",
  c("benchmarks/mirai_runtime_benchmark.R", "benchmark-results.csv")
)
```

The following medians were measured over three repetitions on an Apple M1 Pro
with 32 GB RAM and R 4.6.1. The daemon was warmed before measurement; filesystem
measurements used the operating system's normal warm page cache.

| Workload | Mode | Return/submit (ms) | Event-loop lag (ms) | Total (ms) | Payload (MB) |
|---|---:|---:|---:|---:|---:|
| HLA motif graph, 12,000 receptors | `master` synchronous | 663 | 663 | 663 | 0 |
| HLA motif graph, 12,000 receptors | `mirai` asynchronous | 4 | 4 | 414 | 4.231 |
| Moran's I, 1,200 cells | `master` synchronous | 73 | 73 | 73 | 0 |
| Moran's I, 1,200 cells | `mirai` asynchronous | <1 | <1 | 100 | 0.029 |
| CRB load in a later session | `master` per-session | 121 | 121 | 121 | 0 |
| CRB load in a later session | process cache hit | <1 | <1 | <1 | 0 |
| Linked Views bundle in a later session | `master` per-session | 351 | 351 | 351 | 0 |
| Linked Views bundle in a later session | process cache hit | <1 | <1 | <1 | 0 |

The primary result is the event-loop column. HLA submission reduced measured
main-loop occupancy by about 99.4%; Moran submission returned within the clock's
millisecond resolution. Moran total time increased because daemon dispatch and
serialization dominate this small synthetic case. That is an acceptable
trade-off for an interactive multi-session server; very small work remains a
candidate for synchronous execution if production traces show dispatch costs
outweighing contention.

## Browser startup and first-open latency

`benchmarks/viewer_latency_benchmark.R` drives the unchanged `master` app and
the optimized app through Chrome with `shinytest2`. Each repetition starts a
fresh R process and browser, waits for Data info, allows a fixed 1.5 second
post-paint window, then opens Projection, Linked Views, and Gene Expression. The
Gene Expression measurement includes selecting `MS4A1` and receiving the first
canvas. Run it with two independent checkouts:

```{r browser-benchmark-command, eval=FALSE}
system2(
  "Rscript",
  c(
    "benchmarks/viewer_latency_benchmark.R",
    "--baseline=/path/to/master-checkout",
    "--candidate=.",
    "--reps=3",
    "--output=benchmarks/results/viewer_latency.csv"
  )
)
```

The following are the three-repetition medians recorded on 7 September 2026
using the bundled PBMC demo, Apple M1 Pro, 32 GB RAM, R 4.6.1, and Chrome. Raw
observations are retained in
`benchmarks/results/viewer_latency_2026-09-07.csv`.

| Browser milestone | `master` (ms) | Optimized (ms) | Change |
|---|---:|---:|---:|
| Fresh process to Data info | 3,383 | 3,086 | -8.8% |
| Browser connected to Data info | 1,580.8 | 842.5 | -46.7% |
| Projection first open | 525 | 413 | -21.3% |
| Linked Views first open | 44 | 45 | +2.3% |
| Projection then Linked Views | 569 | 459 | -19.3% |
| Gene Expression first open and first gene canvas | 1,019 | 633 | -37.9% |

The post-paint window is intentional. Once Data info is usable, the optimized
app prepares Linked Views while the browser is otherwise idle and no dedicated
cell-view canvas is active. This keeps the initial response out of the critical
path without moving a roughly 300 ms cold render onto the user's first Linked
Views click. If the user opens Projection or another dedicated cell view before
the warm-up begins, that page keeps ownership of the shared canvas and the
warm-up waits.

Fresh-process time is dominated by starting R and Chrome and therefore barely
changes. The app-specific connected-to-Data-info interval is the relevant
first-paint measure and falls by 46.7%.

# Concurrency and failure semantics

Each interactive workload has a small per-session result cache. When a user
changes a gene, chain, scope, or trajectory before the prior task finishes, the
runtime stops the prior mirai when possible and advances a generation token.
Any late callback from the old generation is ignored. Session termination
cancels its outstanding tasks.

Worker failures are converted to Shiny validation messages rather than escaping
through the server loop. Existing daemons under the configured compute name are
preserved; CerebroNexus only stops a daemon profile it started itself.

# Validation

Focused tests cover configuration, lifecycle ownership, cancellation, stale
result suppression, bounded caches, real out-of-process execution, standalone
app contracts, pure-worker parity, and every workload integration. The full
package test suite remains the final compatibility check:

```{r tests, eval=FALSE}
devtools::load_all(".")
testthat::test_dir(
  "tests/testthat",
  reporter = testthat::default_reporter(),
  stop_on_failure = TRUE
)
```

On 7 September 2026 the final source tree produced `11,319` passing assertions,
zero failures, zero errors, zero warnings, and one environment-dependent skip
in the complete `tests/testthat` run. The browser coverage includes the installed
app form, standalone `inst/app.R`, Data-info-first paint, idle Linked Views
warm-up, Projection, Gene Expression, and immune-repertoire/Clonal UMAP paths.
`node --check inst/viewer/www/cell_views.js` also passed. The benchmark table
above is the median of the 36 raw observations retained in the repository, not
an extrapolation from unit timings.

# Operational interpretation

`mirai` does not create unlimited capacity. The queue remains bounded and two
workers remain two workers. For production, scale Shiny processes for user
isolation and use these daemons to keep each process responsive. Monitor queue
wait, task duration, cancellation rate, payload size, and worker RSS before
raising the worker count or queue memory; more workers can reduce latency while
increasing memory pressure from loaded R packages and copied task inputs.
