---
title: "Trekker positioning evidence and robustness"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Trekker positioning evidence and robustness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# What this module adds

Trekker coordinates are inferred from the spatial barcodes associated with a
nucleus. They are not directly observed image coordinates. CerebroNexus
therefore treats the final position and the evidence supporting that position
as separate data:

```text
Seurat expression and metadata
        +
official Trekker companions
        |
        v
.crb metadata and parsed tables + .trekker/ original files and images
        |
        v
positioning evidence -> active cohort -> robustness analyses
```

The Viewer does more than place nuclei on another scatter plot. It can expose
the vendor-provided evidence for each position, let the user select nuclei by
an evidence range, and recompute downstream summaries for that active cohort.
It does not combine unrelated evidence fields into an uncalibrated confidence
score.

# Explicit Trekker declaration

A Seurat object does not prove how its spatial coordinates were generated.
CerebroNexus marks a dataset as Trekker only when the user explicitly supplies
official Trekker companion data through `trekker_data`.

The Location file is the minimum contract:

| Category | Input | Required | Stored use |
|---|---|---:|---|
| `location` | `*Location_ConfPositionedNuclei*.csv` | yes | authoritative physical coordinates |
| `positioning` | `coords_<sample>.txt` | no | nucleus-level positioning evidence |
| `metrics` | `*summary_metrics*.csv` | no | run QC and positioning funnel |
| `cluster_markers` | `*variable_features_clusters*.csv` | no | official marker evidence |
| `moran` | `*variable_features_spatial_moransi*.txt` | no | official spatial-gene evidence |
| `report` | `*Trekker_Report*.html` | no | unchanged downloadable report |

`positioning_evidence` is accepted as an input alias for `positioning`; the
stored category is always `positioning`. The `cell_bc` values in the positioning
file are aligned to the nuclei in the Location/CRB contract. Missing, duplicate,
or ambiguous barcode matches stop the import.

# Export directly from Seurat

Only one Seurat object is required. Supply the official Location file to create
a valid Trekker CRB, then add whichever optional companions are available:

```{r export-from-seurat}
library(CerebroNexus)

exportFromSeurat(
  object = seurat,
  assay = "RNA",
  slot = "data",
  file = "sample.crb",
  experiment_name = "Trekker sample",
  organism = "mm",
  groups = c("sample", "seurat_clusters"),
  trekker_data = list(
    location = "sample_Location_ConfPositionedNuclei.csv",
    positioning = "coords_sample.txt",
    metrics = "sample_summary_metrics.csv",
    cluster_markers = "sample_variable_features_clusters.csv",
    moran = "sample_variable_features_spatial_moransi.txt",
    report = "sample_Trekker_Report.html"
  )
)
```

Parsed coordinates, evidence, and analysis tables are stored in the CRB.
Original companion files remain byte-for-byte unchanged in the sibling
`sample.trekker/` directory, together with their checksums. Keep the CRB and
this directory together.

# Add Trekker data while building an app

An ordinary CRB can become a Trekker dataset during app creation. This is useful
when the expression export has already finished:

```{r enrich-app}
createShinyApp(
  cerebro_data = c("Trekker sample" = "sample.crb"),
  result_dir = "trekker-app",
  trekker_data = list(
    location = "sample_Location_ConfPositionedNuclei.csv",
    positioning = "coords_sample.txt",
    metrics = "sample_summary_metrics.csv"
  ),
  initial_page = "trekker"
)
```

The source CRB is not modified. `createShinyApp()` enriches a staged copy and
places the original Trekker files in the app's private data directory.

## Conflict and replacement rules

When both the CRB and `createShinyApp()` provide the same category:

- identical content is deduplicated;
- different content raises an error rather than silently overwriting data;
- `positioning`, `metrics`, `cluster_markers`, `moran`, and `report` can be
  replaced only when named explicitly in `trekker_replace`;
- Location cannot be replaced while building an app because it defines the
  authoritative nucleus coordinates. Re-export the CRB to change it.

For one dataset:

```{r replace-positioning}
createShinyApp(
  cerebro_data = c("Trekker sample" = "sample.crb"),
  result_dir = "trekker-app",
  trekker_data = list(positioning = "coords_sample_revised.txt"),
  trekker_replace = "positioning"
)
```

For several datasets, key both arguments by the names in `cerebro_data`.

# Sections and image layers

A Trekker run may contain one or more sections/FOVs. Every nucleus belongs to
exactly one section, and every section may have several ordered image layers.
Image bytes are not embedded in the CRB.

```{r sections-and-images}
exportFromSeurat(
  object = seurat,
  assay = "RNA",
  slot = "data",
  file = "brain.crb",
  experiment_name = "Brain",
  organism = "mm",
  groups = "seurat_clusters",
  trekker_data = list(
    location = "brain_Location_ConfPositionedNuclei.csv"
  ),
  trekker_sections = c(
    cell_a = "slice1",
    cell_b = "slice2"
  ),
  trekker_images = list(
    slice1 = list(DAPI = "slice1-dapi.png", IF = "slice1-if.png"),
    slice2 = list(DAPI = "slice2-dapi.png", IF = "slice2-if.png")
  ),
  trekker_image_settings = list(
    slice1 = list(
      DAPI = list(image_opacity = 0.55, flip_y = TRUE)
    )
  )
)
```

The Viewer provides a multi-select **Layer image** control for the current
section. Each selected layer has independent opacity, offset, scale, rotation,
aspect-ratio, and flip controls under **Settings**. Image layers are hidden in
the transition view because that view interpolates between UMAP and physical
coordinates on a single coordinate system.

`createShinyApp()` accepts the same section, image, and settings structures. If
an existing `(section, image label)` receives different bytes, list that exact
identity under `trekker_image_replace`; equal SHA-256 content is deduplicated.

# Evidence and robustness workflow

## Inspect positioning evidence

When `positioning` is available, the **Overview** tab exposes numeric fields
that are present in the official `coords_<sample>.txt`, including candidate
location counts, spatial-barcode/UMI support, noise fractions, top-location
fractions, and recovery status.

Selecting one nucleus shows its source barcode, supplied evidence fields, final
Location coordinates, section, gene expression, and CRB metadata. Values are
shown as vendor evidence; the Viewer does not reinterpret them as a probability
that the position is correct.

## Create an active cohort

Choose an evidence field and range, then select nuclei inside or outside that
range. The same active cohort is sent back to the main physical map and used by:

- active-cohort RNA QC;
- neighbourhood enrichment and composition;
- cluster spatial profiles;
- local neighbourhood maps;
- UMAP-layout versus physical-layout neighbour overlap.

Clearing the evidence filter restores analysis of the complete current section.
The active cohort can also be downloaded with barcode, section, Location
coordinates, and CRB metadata.

This supports a direct robustness question: do the visible marker, QC, and
neighbourhood patterns remain after nuclei with weak or atypical positioning
evidence are excluded?

# Viewer outputs and interpretation

The Trekker page combines four evidence layers:

1. **Physical context**: Location map, optional UMAP pair/transition, external
   image layers, metadata and expression colouring.
2. **Positioning and run QC**: per-nucleus evidence, positioning funnel, read and
   barcode retention, depth, ambiguity, and recovery summaries.
3. **Biological evidence**: official cluster markers, official Moran results,
   gene maps, and multi-gene summaries.
4. **Exploratory robustness**: cohort-aware neighbourhood and section
   comparisons computed by the Viewer.

The **Files** tab reports provenance and checksums and lets users download the
original companions and image layers. Official HTML is retained for download;
it is not executed inside the Viewer.

## Scale limits

Display and exploratory neighbourhood calculations are bounded for large runs:

- physical expression and QC previews use at most 8,000 deterministically
  sampled nuclei;
- neighbourhood summaries use at most 1,000 deterministically sampled nuclei;
- neighbourhood empirical P values use 199 fixed-seed label permutations and
  BH correction.

These calculations are interactive summaries, not a replacement for a
pre-registered spatial statistics workflow. In particular, the UMAP-layout
overlap compares two two-dimensional layouts; it must not be described as
agreement between physical space and the original high-dimensional molecular
space.

# Bundled demo

The bundled Mouse Brain TrekkerU C demo is a size-limited subset of official
Location, run-QC, marker, Moran, Seurat expression, and metadata files. The
public example bundle does not contain `coords_<sample>.txt`, so the demo does
not fabricate positioning evidence.

Its two sections and four small SVG layers exist only to exercise the
multi-section and multi-layer interface. They are labelled as synthetic UI
overlays and must not be interpreted as measured tissue images or independent
biological sections.

# Further reading

- [Data integrity from Seurat to a self-contained Cerebro app](data_integrity_contracts.html)
- [Create a self-contained Shiny app](create_a_self_contained_shiny_app.html)
- [Spatial transcriptomics analysis](spatial_analysis.html)
